---
title: "NE Nigeria Climate and Health/Nutrition Analysis"
output: github_document
---
<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = F,
  message=F,
  results="hide"
)
```


<!-- badges: start -->
<!-- badges: end -->
Repository to store analysis of climate and health data in NE in Nigeria

```{r, echo=FALSE,message=F}
library(tidyverse)
library(dygraphs)
library(xts)

invisible(purrr::map(list.files(here::here("R"),full.names = T),.f = source))
```

```{r, echo=F}

awd <- readxl::read_excel(here::here("data/Borno_EWARS_awd_under5_2017_2021.xlsx")) |> 
  rename(ADM2_EN = `...1`)
  
nga_admins <- invisible(load_nga_admins())

ne_lgas <- nga_admins$lga
awd <- awd |> 
  mutate(
    ADM2_EN= str_replace_all(string = ADM2_EN,c("Askira Uba"="Askira/Uba","Kala Balge"="Kala/Balge" ))
  )

awd_long <- awd |> 
  pivot_longer(-1,names_to = "date",values_to = "cases") |> 
  left_join(ne_lgas |> select(ADM1_EN,ADM2_EN)) |> 
  select(ADM1_EN,ADM2_EN,everything())

awd_long_mj <- awd_long %>% 
  filter(ADM2_EN %in% c("Maiduguri","Jere"))

# add year-wk - double checked there is only 1 week collected per year per location
awd_long_mj <- awd_long_mj %>%
  mutate(
    yr_wk=paste0(lubridate::year(date),"_",lubridate::week(date))
  )

awd_long_by_wk <- awd_long_mj %>% 
  mutate(
    date= lubridate::ymd(date),
    cases= as.numeric(cases)
  ) %>% 
  group_by(date) %>% 
  summarise(
    value=sum(cases,na.rm=T),.groups = "drop"
  ) %>% 
  mutate(
    yr_wk=paste0(lubridate::year(date),"_",lubridate::week(date)),
    parameter="awd_cases"
  )


```

```{r}

smap <- read_rds(file = here::here("data/SMAP_soil_moisture_3day_maidu_jere.rds"))

chirps_daily_long <- read_rds(here::here("data/chirps_daily_maidu_jere_diss_gte2016.rds"))
chirps_daily_long <- chirps_daily_long %>% 
  group_by(yr=lubridate::year(date)) %>% 
  mutate(
    cumulative_precip=cumsum(value)
  ) %>% 
  ungroup()
  
smap_chirps_awd_long <- bind_rows(chirps_daily_long %>% ungroup() %>% 
  select(-cumulative_precip,-yr),
  smap,awd_long_by_wk %>% 
    select(-yr_wk)
    )


smap_chirps_awd_wide<- smap_chirps_awd_long %>% pivot_wider(names_from = parameter,values_from = value)

dat_wide_filt <- smap_chirps_awd_wide %>% 
  select(date,precip,ssm,susm,awd_cases) %>% 
  mutate(awd_cases=awd_cases/100) %>% 
  filter(date>="2016-01-01")

dat_xts<-xts(dat_wide_filt,order.by =  dat_wide_filt$date)

precip_awd_dygraph<-dygraph(dat_xts) %>% 
  dySeries("precip", label = "precip", axis=('y'),color="#74add1") %>%
  dySeries("ssm", label = "surface moisture" ,axis=('y'),color="#a6d96a") %>% 
  dySeries("susm", label = "sub-surface moisture", axis=('y'),color="#8c510a")  %>% 
  dySeries("awd_cases", label = "awd_cases", axis=('y'),color="red")  %>% 

  dyRangeSelector(dateWindow = c(min(dat_xts$date),
                                 max(dat_xts$date))) %>% 
  dyOptions(connectSeparatedPoints = TRUE)


```



## Overview  of Data
- Below is a graph showing:
  + precipitation
  + surface soil moisture
  + sub-surface soil moisture
  + AWD cases/100
  
```{r, results='markup'}

htmltools::tags$iframe(
  src=file.path(here::here("README_files/figure-gfm"), "awd_cases_precip_moisture.html"),
  width="100%",
  height="600",
  scrolling="no",
  seamless="seamless",
  frameBorder="0"
)

```




## Establishing Environmental Thresholds

- A prominent peak occurs shortly after the start of the rainy season. 
- Rainfall generally declines after this initial peak
- Below we have graphed cumulative yearly rainfall against AWD caseload
  + we see that AWD caseloads tend to peak in a narrow band of cumulative rainfall
```{r,echo=F,results="asis"}


# chirps_daily_long <- chirps_daily_long %>% 
#   group_by(yr=lubridate::year(date)) %>% 
#   mutate(
#     cumulative_precip=cumsum(value)
#   ) 

yrly_max_case_weeks <- awd_long_by_wk %>% 
  group_by(yr=lubridate::year(date)) %>%
  slice_max(order_by =value, n = 1 ) %>% 
  # filter(value==max(value)) %>% 
  mutate( 
    yrmo = paste0(yr,"_",lubridate::month(date))
  )
awd_first_peaks <- awd_long_by_wk %>% 
  filter(
    date %in%c(yrly_max_case_weeks$date[1:3],
               "2020-07-19",
               "2021-07-11")
  )

chirps_daily_long%>% 
  ggplot(aes(x=date,cumulative_precip))+
  geom_line(color="#abd9e9",lwd=1)+
  geom_line(data= awd_long_by_wk,
            aes(x=date,y=value/2),
            lwd=1,color="#fdae61",alpha=0.75)+
 
  labs(y="Yearly cumulative precipitation (mm)")+
  # drawing lines roughly where the rains start each year
  # geom_vline(xintercept=lubridate::ymd("2017-05-01"))+
  # geom_vline(xintercept=lubridate::ymd("2018-04-01"))+
  # geom_vline(xintercept=lubridate::ymd("2019-05-01"))+
  # geom_vline(xintercept=lubridate::ymd("2020-05-01"))+
  # geom_vline(xintercept=lubridate::ymd("2021-05-01"))+
  geom_vline(xintercept=lubridate::ymd("2017-07-09"),color="#f46d43",lwd=1)+
  geom_vline(xintercept=lubridate::ymd("2018-07-15"),color="#f46d43",lwd=1)+
  geom_vline(xintercept=lubridate::ymd("2019-06-30"),color="#f46d43",lwd=1)+
  geom_vline(xintercept=lubridate::ymd("2020-07-19"),color="#f46d43",lwd=1)+
  # geom_vline(xintercept=lubridate::ymd("2020-08-30"),)+
  geom_vline(xintercept=lubridate::ymd("2021-07-11"),color="#f46d43",lwd=1)+
  geom_hline(yintercept=127,color="#4575b4",lwd=1, linetype="dashed")+
  geom_hline(yintercept=184,color="#4575b4",lwd=1, linetype="dashed")+
  # geom_vline(xintercept=lubridate::ymd("2021-08-29"))+
  geom_point(data=awd_first_peaks, aes(x= date, y= value/2),fill="#fdae61", size= 2,pch=21)+
  
  # seems like cases are spiking between 70-300 mm of rain accumulation/
  # first half of the rainy season
  scale_y_continuous(
    sec.axis= sec_axis(trans = ~.x*2,name = "AWD Cases")
  )+
  scale_x_date(
    minor_breaks = "1 month",
    breaks= "1 month",
    # major="year",
    labels=scales::date_format("%b%y")
  )+
  theme_bw()+
  theme(
    axis.text.x = element_text(angle=90,size = 6)
  )+
  annotate(
    "text", label = "127 mm",
    x = lubridate::ymd("2015-10-01"), y = 115, size = 3, colour = "#4575b4"
  )+
  annotate(
    "text", label = "186 mm",
    x = lubridate::ymd("2015-10-01"), y = 200, size = 3, colour = "#4575b4"
  )

```