---
title: "monthly_env_data_wards"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries & functions



```{r libraries}
library(tidyverse)
library(rgee)
library(sf)
invisible(purrr::map(list.files(here::here("R"),full.names = T),.f = source))
```

Now that we have a better understanding of data and limitations it makes sense to re-think how the data is extracted from GEE. Once it is extracted it should be cleanly compiled into several long format data.frames which can be stored as a list and saved as an `rds` file.

- OTP data 
  + All 3 states: Yobe, Maiduguri, Adamawa at LGA level (still some LGA cleaning required)
  + No ward level data
- AWD Data
  + Just Borno State at LGA & Ward level. Ward data is too dirty to be used for entire state.
  + Majority of data in Maiduguri & Jere. We have had those wards cleaned to acceptable level
  

  
## Data


```{r load_data}

ee_Initialize()

# 16 day ndvi composites
ndvi_dat_src <- "MODIS/006/MOD13Q1"
ndvi <- ee$ImageCollection(ndvi_dat_src)

# ee_get_date_ic(ndvi)

soil_moisture_src <- "NASA_USDA/HSL/SMAP10KM_soil_moisture"
soil_moisture <- ee$ImageCollection(soil_moisture_src)

chirps_src <- "UCSB-CHG/CHIRPS/DAILY"
chirps <- ee$ImageCollection(chirps_src)

# awd <- readxl::read_excel(here::here("data/Borno_EWARS_awd_under5_2017_2021.xlsx")) |> 
#   rename(ADM2_EN = `...1`)

muac <- read_rds(file = here::here("data/muac_201701_202111.rds"))
admins <- load_nga_admins()

ward <- admins$ward

# wards of interest
woi <- ward %>% 
  filter(ADM2_EN %in% c("Maiduguri","Jere"))

```


# get precip

```{r}

monthly_precip <- yr_mo_composite_stat_ic(dat = chirps,
                                          month_range = c(1,12),
                                          year_range = c(1981,2021),
                                          stat = "sum")



monthly_precip_renamed <- monthly_precip$ 
  # CHIRPS release is delayed by a month or 2 so you need to pre-filter
  # or face difficulty later
  filterDate("1981-01-01","2021-11-30") |> 
  ee$ImageCollection$map(
    function(x) {
      date <- ee$Date(x$get("system:time_start"))$format('YYYY_MM_dd')
      name <- ee$String$cat("pr_", date)
      x$select("precipitation")$rename(name)
    })


system.time(
monthly_precip_ward <- ee_extract(x = monthly_precip_renamed,y=woi["ADM3_PCODE"],scale=5500,fun= ee$Reducer$mean(),sf=T)
)

monthly_precip_ward_long <- monthly_precip_ward |> 
       st_drop_geometry() |> 
  pivot_longer(-1) |> 
      mutate(
        parameter="precip",
        date= str_remove(string = name, pattern = ".*pr_") |> 
          str_replace_all("_","-") |> lubridate::ymd()

  ) 
  


```

## Soil Moisure
```{r, eval=F}
soil_moisture %>% ee_get_date_ic() %>% pull(time_start) %>% range()
# soil_moisture
soil_moisture$first()$bandNames()$getInfo()
soil_moisture$select("ssm")$first() %>% ee_print()


monthly_soil_moisture <- yr_mo_composite_stat_ic(dat = soil_moisture$
                                         select("ssm","susm","smp"),

                                          month_range = c(1,12),
                                          year_range = c(2015,2021),
                                          stat = "mean")







monthly_soil_moisture <- monthly_soil_moisture$
  filterDate("2015-04-02","2022-01-10") 

soil_moisture_3_day <- soil_moisture$
  filterDate("2015-04-02","2022-01-10") 

monthly_soil_moisture_renamed <- monthly_soil_moisture |> 

  ee$ImageCollection$map(
    function(x) {
      date <- ee$Date(x$get("system:time_start"))$format('YYYY_MM_dd')
      # band_names <- x$bandNames()$getInfo()
      # band_names_date <- 
      ssm_date <- ee$String$cat("ssm_", date)
      susm_date <- ee$String$cat("susm_", date)
      smp_date <- ee$String$cat("smp_", date)
      x$rename(ssm_date,susm_date,smp_date)
      
    })


soil_moisture_3_day_renamed <- soil_moisture_3_day$
  select("ssm","susm","smp") %>% 

  ee$ImageCollection$map(
    function(x) {
      date <- ee$Date(x$get("system:time_start"))$format('YYYY_MM_dd')
      # band_names <- x$bandNames()$getInfo()
      # band_names_date <- 
      ssm_date <- ee$String$cat("ssm_", date)
      susm_date <- ee$String$cat("susm_", date)
      smp_date <- ee$String$cat("smp_", date)
      x$rename(ssm_date,susm_date,smp_date)
      
    })



# should look up if this resolution (12.7 km) should be used rather than the 10 km provided by GEE
soil_moisture$first()$projection()$nominalScale()$getInfo()


woi_diss <-  woi %>% 
  summarise() %>% 
  st_simplify(dTolerance =  0.7)

system.time( #24 s
soil_moisture_3_day_mj <- ee_extract(x = soil_moisture_3_day_renamed,
                                         y=woi_diss,
                                         scale=10000,
                                         fun= ee$Reducer$mean(),
                                         sf=T)
)


system.time( #24 s
monthly_soil_moisture_ward <- ee_extract(x = monthly_soil_moisture_renamed,y=woi["ADM3_PCODE"],scale=10000,fun= ee$Reducer$mean(),sf=T)
)

monthly_soil_moisture_ward_long <- monthly_soil_moisture_ward |> 
       st_drop_geometry() |> 
      pivot_longer(-1) |> 
      mutate(
        parameter= str_extract(string = name,pattern = "smp|ssm|susm"),
        date= str_remove(string = name, pattern = ".*sm_|.*mp_") |> 
          str_replace_all("_","-") |> lubridate::ymd()
  )
  
soil_moisture_3_day_mj_long <- soil_moisture_3_day_mj |> 
       st_drop_geometry() |> 
      pivot_longer(-1) |> 
      mutate(
        parameter= str_extract(string = name,pattern = "smp|ssm|susm"),
        date= str_remove(string = name, pattern = ".*sm_|.*mp_") |> 
          str_replace_all("_","-") |> lubridate::ymd()
  )

# soil_moisture_3_day_mj_long %>% 
#     select(date,parameter,value) %>% 
#   write_rds("SMAP_soil_moisture_3day_maidu_jere.rds")


```

## NDVI
- need to work in VCI
<!-- https://journals.ametsoc.org/view/journals/clim/23/3/2009jcli2900.1.xml -->
  + VCI = (NDVI-NDVImin)/(NDVImax-NDVImin)
  +VHI

```{r}

ndvi_rescaled <- ndvi$
  select("NDVI")$
  map(
    ee_utils_pyfunc(
      function(x){x$
          multiply(0.0001)$
          copyProperties(x,x$propertyNames())
          }
    )
  )




ndvi_rescaled_monthly <- yr_mo_composite_stat_ic(dat =ndvi_rescaled,
                                                 month_range = c(1,12),
                                                 year_range = c(2016,2021),
                                                 stat = "mean")



ndvi_rescaled_monthly <- ndvi_rescaled_monthly$select("NDVI")
ndvi_rescaled_monthly %>% ee_print()
month_sequence <- ee$List$sequence(1,12)
year_sequence <- ee$List$sequence(2016,2021)

ndvi_rescaled_monthly_historical_mean <- month_sequence$map(
   ee_utils_pyfunc(
     function(m) {
       ndvi_mean <- ndvi_rescaled$filter(ee$Filter$calendarRange(2000, 2010, 'year'))$
         filter(ee$Filter$calendarRange(m, m, 'month'))$
         mean()$
         rename("NDVI_mean")$
         set("system:time_start", ee$Date$millis(ee$Date$fromYMD(2000,m,1)) )
       
       ndvi_std <- ndvi_rescaled$filter(ee$Filter$calendarRange(2000, 2010, 'year'))$
         filter(ee$Filter$calendarRange(m, m, 'month'))$
         reduce(ee$Reducer$stdDev())$
         rename("NDVI_std")$
         set("system:time_start", ee$Date$millis(ee$Date$fromYMD(2000,m,1)) )
       
       
       ndvi_mean$addBands(ndvi_std)
     }

)
)


# try simplifying above
ndvi_rescaled_monthly_historical_mean <- month_sequence$map(
   ee_utils_pyfunc(
     function(m) {
       ndvi_tfilt <- ndvi_rescaled$filter(ee$Filter$calendarRange(2000, 2010, 'year'))$
         filter(ee$Filter$calendarRange(m, m, 'month'))
       
       
       ndvi_mean <- ndvi_tfilt$
         mean()$
         rename("NDVI_mean")$
         set("system:time_start", ee$Date$millis(ee$Date$fromYMD(2000,m,1)) )
       
       ndvi_min <- ndvi_tfilt$
         min()$
         rename("NDVI_min")$
         set("system:time_start", ee$Date$millis(ee$Date$fromYMD(2000,m,1)) )
       
       ndvi_max <- ndvi_tfilt$
         max()$
         rename("NDVI_max")$
         set("system:time_start", ee$Date$millis(ee$Date$fromYMD(2000,m,1)) )
       
       ndvi_std <-ndvi_tfilt$
         reduce(ee$Reducer$stdDev())$
         rename("NDVI_std")$
         set("system:time_start", ee$Date$millis(ee$Date$fromYMD(2000,m,1)) )
       
       
       
       ndvi_mean$addBands(ndvi_min)$addBands(ndvi_max)$addBands(ndvi_std)
     }

)
)

ndvi_rescaled_monthly_historical_mean <- ee$ImageCollection$fromImages(ndvi_rescaled_monthly_historical_mean$flatten()) 

ndvi_rescaled_monthly_historical_mean %>% ee_print()
ndvi_rescaled_monthly_historical_mean %>% ee_get_date_ic() %>% pull(time_start) %>% range()
ndvi_rescaled_monthly %>% ee_get_date_ic() %>% pull(time_start) %>% range()


ndvi_rescaled_monthly$first()$bandNames()$getInfo()
ndvi_rescaled_monthly_historical_mean$first()$bandNames()$getInfo()



recent_vs_historical_ndvi <- year_sequence$map(
  ee_utils_pyfunc(
     function(y) {
       month_sequence$map(
         ee_utils_pyfunc(
           function(m){
             recent_yrmo<- ee$Image(ndvi_rescaled_monthly$
              filter(ee$Filter$calendarRange(y, y, 'year'))$
              filter(ee$Filter$calendarRange(m, m, 'month'))$
                first()
             )
             
             hist_mo <- ee$Image(ndvi_rescaled_monthly_historical_mean$
                                   filter(ee$Filter$calendarRange(2000, 2000, 'year'))$
                                   filter(ee$Filter$calendarRange(m, m, 'month'))$
                                   first()
               )
             
            
            historical_min <- hist_mo$select("NDVI_min")
            historical_max <- hist_mo$select("NDVI_max")
            
            NDVI_diff <- recent_yrmo$select("NDVI")$
              subtract(hist_mo$select("NDVI_mean"))$
              rename("NDVI_diff")
            
            NDVI_z <- NDVI_diff$select("NDVI_diff")$
              divide(hist_mo$select("NDVI_std"))$
              rename("NDVI_Z")
            
            # max_diff_min <- - historical_max$
            #   subtract(historical_min)
            
            max_diff_min <-  hist_mo$select("NDVI_max")$
              subtract( hist_mo$select("NDVI_min"))$
              rename("max_diff_min")
          
            NDVI_VCI_diff <- recent_yrmo$select("NDVI")$
              subtract(hist_mo$select("NDVI_min"))$
              rename("VCI_NDVI_diff")
            # 
            NDVI_VCI <- NDVI_VCI_diff$
              select("VCI_NDVI_diff")$
              subtract(max_diff_min$select("max_diff_min"))$
              rename("VCI_NDVI")
            
            recent_yrmo$
              addBands(NDVI_diff)$
              addBands(NDVI_z)$
              addBands(NDVI_VCI)$
              set("system:time_start",ee$Date$millis(ee$Date$fromYMD(y,m,1)))
           }
         )
       )
       
     }
  
)
)



recent_vs_historical_ndvi_ic <- ee$ImageCollection$fromImages(recent_vs_historical_ndvi$flatten())
recent_vs_historical_ndvi_ic %>% ee_print()
recent_vs_historical_ndvi_ic$first()$bandNames()$getInfo()


recent_vs_historical_ndvi_ic_renamed <- recent_vs_historical_ndvi_ic |> 
  ee$ImageCollection$map(
    function(x) {
      date <- ee$Date(x$get("system:time_start"))$format('YYYY_MM_dd')
      # name <- ee$String$cat("pr_", date)
      # x$select("NDVI")$rename(name)
      
      ndvi_bandname <- ee$String$cat("NDVI_", date)
      diff_ndvi_bandname <- ee$String$cat("diff_NDVI_", date)
      z_ndvi_bandname <- ee$String$cat("z_NDVI_", date)
      x$rename(ndvi_bandname,diff_ndvi_bandname,z_ndvi_bandname)
    }
  )




system.time( #36 s
monthly_ndvi_params <- ee_extract(x = recent_vs_historical_ndvi_ic_renamed,
                           y=woi["ADM3_PCODE"],
                           scale=250, # dont fogrget to set scale properly
                           fun= ee$Reducer$mean(),sf=T)
)

monthly_ndvi_ward_long <- monthly_ndvi_params |> 
       st_drop_geometry() |> 
      pivot_longer(-1) |> 
      mutate(
        parameter= str_extract(string = name,pattern = "diff_NDVI|_NDVI_|z_NDVI") |> 
          str_replace_all(pattern = "_NDVI_","NDVI"),
        date= str_remove(string = name, pattern = ".*NDVI_") |> 
          str_replace_all("_","-") |> lubridate::ymd()
  )

admin_dissolved <- admins$state %>% 
  summarise() %>% 
  mutate(admin="overall")
monthly_ndvi_params_overall <- ee_extract(x = recent_vs_historical_ndvi_ic_renamed,
                           y=admin_dissolved["admin"],
                           scale=250, # dont fogrget to set scale properly
                           fun= ee$Reducer$mean(),sf=T)


monthly_ndvi_overall_long <- monthly_ndvi_params_overall |> 
       st_drop_geometry() |> 
      pivot_longer(-1) |> 
      mutate(
        parameter= str_extract(string = name,pattern = "diff_NDVI|_NDVI_|z_NDVI") |> 
          str_replace_all(pattern = "_NDVI_","NDVI"),
        date= str_remove(string = name, pattern = ".*NDVI_") |> 
          str_replace_all("_","-") |> lubridate::ymd()
  )



monthly_ndvi_overall_long %>% 
  filter(parameter!="NDVI") %>% 
  ggplot(aes(x=date, y= value))+
   scale_x_date(
    minor_breaks = "1 month",
    breaks= "1 year",
    # major="year",
    labels=scales::date_format("%b%y")
  )+
  # scale_x_date()+
  geom_line()+
  geom_vline(xintercept=lubridate::ymd("2016-03-01"),linetype='dashed')+
  geom_vline(xintercept=lubridate::ymd("2017-03-01"),linetype='dashed')+
  geom_hline(yintercept=0,color="red")+
  facet_wrap(~parameter,nrow=2, scales = "free_y")+
  theme_bw()
  


```


## Compile RS ward level

```{r}
env_params_ward<- list(monthly_precip_ward_long,
                       monthly_ndvi_ward_long,
                       monthly_soil_moisture_ward_long) %>% 
  map_dfr(~.x %>% 
            select(ADM3_PCODE, value, parameter, date))


env_params_ward %>% 
  count(parameter)

env_params_ward %>% 
  filter(date>="2016-01-01") %>% 
  count(parameter)

env_params_ward_gte2016 <- env_params_ward %>% 
    filter(date>="2016-01-01") 

env_params_ward_gte2016 %>% 
  count(parameter)

env_params_ward_gte2016_labelled <- env_params_ward_gte2016 %>% 
  left_join(woi %>% 
              st_drop_geometry() %>% 
              select(ADM3_PCODE,ADM3_EN, ADM2_EN,ADM2_PCODE,ADM1_EN)) %>% 
  select(date,matches("^ADM"), parameter,everything())
  
```


```{r}


env_params_ward_gte2016_labelled %>% 
  filter(str_detect(parameter,"NDVI")) %>% 
  filter(parameter %in% c("diff_NDVI","z_NDVI")) %>% 
  filter(ADM3_EN=="Alau") %>% 
  ggplot(aes(x= date, y= value))+
  geom_line()+
  scale_x_date(
    minor_breaks = "1 month",
    breaks= "1 year",
    # major="year",
    labels=scales::date_format("%b")
  )+
  geom_hline(yintercept=0, color="red")+
  facet_wrap(~parameter,nrow=3,scales = "free_y")+
  theme(
    axis.text.x = element_text(angle=90)
  )

# 
# env_params_ward_gte2016_labelled %>% 
#   write_csv("nga_env_params_by_ward_maidu_jere.csv")

```



```{r}
muac_monthly_ward <- muac %>% 
  group_by(State,LGA,Ward,yr,mo) %>% 
  summarise(
    new_otp_admissions = sum(new_otp_admissions,na.rm=T)
  )


# there are some records with LGA missing, but ward present- should recode
ward_lga_lookup<- admins$ward$ADM2_EN |> 
  set_names(admins$ward$ADM3_EN)



muac_cleaned <- muac_monthly_ward %>% 
  ungroup() %>% 
  mutate(
    # across(.cols= matches("^muac_|^new_"), ~replace_na(.x,0)),
    across(.cols = c("State","LGA","Ward"),~str_to_title(.x)),
    mo= str_to_title(mo),
    date= paste0(yr,"-",mo,"-01"),
    date_dt= lubridate::ymd(date),
    LGA= str_replace_all(LGA, c("Askira-Uba"="Askira/Uba",
                                "Postiskum" ="Potiskum" ,
                                "Gaidam" ="Geidam",
                                "Tarmuwa" ="Tarmua" )),
    # fix these na LGAs that have wards
    LGA = if_else(is.na(LGA),recode(Ward, !!!ward_lga_lookup),LGA)
    
    )
ward %>% 
  mutate(
    ward_name= str_to_title(ADM3_EN)
  ) %>% 
  filter(str_detect(ward_name,"Gaji"))
woi %>% 
  pull(ADM3_EN)
muac_cleaned %>% 
  # 7 records of Gajigana in Maiduguri, but admin boundaries say it should be in Magumeri...
  mutate(
    LGA2 = if_else(str_detect(Ward,"Gaji"),"Magumeri",LGA)
  ) %>% 
  filter(str_detect(Ward,"Gaji"))
  filter(State=="Borno") %>% 
  filter(LGA%in% c("Maiduguri","Jere")) %>% 
  
  count(Ward) %>% 
  mutate(
    ward2= str_to_title(str_squish(Ward)),
    ward3 = case_when(
      ward2 %in% c("Bolori Ii",
                   "Bolori Ii",
                   "Bolori-Ii",
                   "Bolori 2",
                   "Bolori 2a&2c",
                   "Bolori 2b&D")~"Bolori II",
      # apparently there is teachers village in both bolori 1 & 2 - did both
      str_detect(ward2,"^Bolori.*|^Teacher.*|^Pompo.*")~"Bolori I",
      str_detect(ward2,"^Bulab.+in$|^Bulab.+")~"Bulablin",
      str_detect(ward2,"^Dala.+")~"Dala",
      str_detect(ward2,"^Dus[um].+|^Muna_eth.*")~"Dusuma",
      str_detect(ward2,"^G[ao]mboru.+")~"Gamboru",
      str_detect(ward2,"^Gomari.*|Gangamari")~"Gomari",
      ward2 %in% c("Gwange","Gwange I","Gwange 1&2","Gwange-I","Gwnge 1")~"Gwange I",
      ward2 %in% c("Gwange Ii")~"Gwange II",
      ward2 %in% c("Gwange Iii","Gwange 3")~"Gwange III",
      ward2 %in% c("Shehuri","Shehuri North","Shehuru North")~"Shehuri North",
      str_detect(ward2,"^Shuwari.*|Shok")~"Shehuri North",
      str_detect(ward2,"^Maisandari.*")~ "Maisandari" ,
      str_detect(ward2,"^Lam[iu]sula*")~ "Lamisula" ,
      str_detect(ward2,"^Musari*|Maimusari.*")~ "Maimusari" ,
      str_detect(ward2,"^Mashamari*")~ "Mashamari" ,
      
      #,https://reliefweb.int/sites/reliefweb.int/files/resources/referral_pathway_mmc_jere_final_24.03.17.pdf
      str_detect(ward2,"^Gw[oa]zar*")~ "Dusuma" ,
      str_detect(ward2,"^Sabon.*|^Bale.*")~ "Galtimari",
      # Gajigana  - there is a ward called Gajigana in Magumeri... maybe it should b
      TRUE~ ward2
    )
  ) %>% 
  count(ward3) %>% 
  print(n=460)
  
```









