---
title: "North East Nigeria AWD, Malnutrition and Climate"
output: 
  bookdown::html_document2:
    toc: true
    toc_depth: 3
    toc_float: true
    # number_sections: true
    self_contained: yes
    # css: !expr here::here("hnap_style.css")
    highlight: pygments
    code_folding: hide
    theme:
      bg: "#FFFFFF"
      fg: "#1E1E1E"
      primary: "#ee5859"
      secondary: "#58c6ee"
# runtime: shiny

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


# thematic::thematic_rmd()
# thematic::thematic_shiny()
# thematic::thematic_on(
#   bg = "#F4F4E2",
#   fg = "#4c4c4c",
#   accent = "#741336",
#   # font = font_spec("Special Elite", scale = 0.5),
#   # qualitative = paletteer::paletteer_d("dutchmasters::milkmaid"),
#   # sequential = sequential_gradient(0.5, 0.75)
# )
# bslib::bs_themer()


```


```{r,message=F,warning=F}
# libs/functions
library(sf)
library(tidyverse)
library(dygraphs)
library(xts)
library(lubridate)
library(glue)
library(thematic)
invisible(purrr::map(list.files(here::here("R"),full.names = T),.f = source))

# "cleaned" data
awd <- invisible(load_awd(lgas = NULL))
monthly_precip_aggs <- read_rds(here::here("data/monthly_precip_aggs.rds"))
smap <- read_rds(file = here::here("data/SMAP_soil_moisture_3day_maidu_jere.rds"))

ndvi <- read_rds(here::here("data/monthly_ndvi_stats_2016_2021.rds"))
otp <-  read_rds(here::here("data/20220121_MUAC_w_maidu_jere_wards_harmonized.rds"))

pop <-  read_csv(here::here("data/pop_lga_msna2019_sf.csv"))
admins$lga |> 
  st_drop_geometry() |> 
  select(matches("^ADM\\d_[PE]"))

```




```{r}
# chunk to explore ACLED data, can remove once i know what to use

library(leaflet)
acled <- read_csv(here::here("data/2015-01-01-2022-02-01--Nigeria_ACLED.csv"))
acled %>% janitor::tabyl(admin1) %>% arrange(desc(n))
acled_sf <- st_as_sf(acled, coords=c("longitude","latitude"),crs=4326)
admins <- load_nga_admins()
ne_states<- admins$state %>% 
  filter(ADM1_EN %in% c("Borno","Yobe","Adamawa"))

ne_wards <- admins$ward %>% 
  filter(ADM1_EN %in% c("Borno","Yobe","Adamawa"))
acled_ne_j<- st_join(acled_sf, ne_wards)
acled_ne_j %>% glimpse()
acled_ne_j <- acled_ne_j %>% 
  filter(!is.na(ADM1_EN)) 

acled_ne_violent <- acled_ne_j %>% 
  filter(str_detect(tolower(event_type),"violence|battle")) 


acled_ne_violent %>% 
  count(sub_event_type)

acled_ne_violent %>% glimpse()

acled_ne_simp <- acled_ne_violent %>% 
  mutate(
    date= dmy(event_date)
  ) %>% 
  select(date, fatalities, event_type,matches("^admin.+|^ADM.+"))


```

# Intro

This page is meant to illustrate research progress and preliminary findings. Full code can be found on the [GitHub repository](https://github.com/zackarno/NE_Nigeria_climate_health). This page contains visualizations, code, and some comments/questions. You can view the code by clicking on the blue `code` buttons. 


# Objective & Motivation

- asdf
- asdfa 
- adsfa


# Summary Statistics/Data Set Descriptions

## Acute Watery Diarhhea (AWD)

- The AWD case-load data was only obtained for Borno State.
- Time interval: weekly
- Data cleaning code can be found in Appendix (section: \@ref(awdCleaning))


Figure \@ref(fig:awdLGA-plot) shows that the majority of AWD case-load data is located in Maiduguri and Jere LGAs.  

```{r awdLGA-plot, fig.cap="Total AWD cases by LGA"}


plot_awd_by_lga <- awd |> 
  group_by(ADM2_EN) |> 
  summarise(
    cases_per_lga= sum(as.numeric(cases)),.groups = "drop"
  ) |> 
  mutate(
    pct_cases= cases_per_lga/sum(cases_per_lga)
  ) |> 
  ggplot(aes(x=reorder(ADM2_EN,cases_per_lga),y=cases_per_lga))+
  geom_bar(stat="identity")+ 
  labs(x="LGA",y="Recorded Cases")+
  coord_flip()


plot_awd_by_lga
```



## Malnutrition (OTP)

- Source: *[Nutrition Cluster 5Ws](https://www.humanitarianresponse.info/en/operations/nigeria/nutrition/documents/document-type/3w-4w)* . 
- Time interval : monthly
- Cleaning/manipulation: See Appendix (Section: \@ref(malnutritionCleaning)).

```{r}

# a little extra cleaning required
otp_cleaned <- otp |> 
  select(date=date_dt, State, LGA=LGA_crct,Ward=ward_crct,new_otp_admissions,muac_total_screened) |> 
  # a little additional cleaning. 
  # assume if there are records for muac screenings then facility was open and NA for OTP admissions = 0
  mutate(new_otp_admissions=if_else(!is.na(muac_total_screened),replace_na(new_otp_admissions,0),new_otp_admissions)) |> 
  filter(!is.na(date)) |> 
  filter(date>"2010-01-01") # weird outlier removal


```

## Remote Sensing variables

Variables of interest were:

- Precipitation
- NDVI
- Soil Moisture

Details regarding the code to aggregate/extract the data using `GEE` & `rgee` can be found in the Appendix (Section: \@ref(climateDataExtraction))



# Temporal View


## OTP & AWD Cases Over Time

Figure \@ref(fig:otpState-plot) show OTP admissions by State. We can see many more records present in Borno State.

```{r otpState-plot, fig.cap="New OTP admissions over time by State (NE NGA)"}

otp_by_state <- otp_cleaned|> 
  group_by(State,date,.drop=F ) |> 
  summarise(
    cases=sum(new_otp_admissions,na.rm=T),.groups = "drop"
  ) 
plot_otp_by_state <- otp_by_state |> 
  ggplot(aes(x=date,y=cases,color=State,group=State))+
  geom_line()+
    scale_y_continuous(labels = scales::comma)+
  scale_x_date(
    # minor_breaks = "months", 
    breaks= "1 month",
    # major="year",
    labels=scales::date_format("%b%y")
    )+
  labs(x= "date", title= "Monthly OTP Caseload by State (NE Nigeria)")+
  theme_bw()+
  theme(
    axis.text.x = element_text(angle=90),
    axis.title.x = element_blank()
  )


plot_otp_by_state


```





In Figure \@ref(fig:awdOtpBorno-plot) we plot both AWD and OTP admissions in Borno together over time. They appear to be related. This relationship is further explored in Figure \@ref(fig:awdVsOtpBorno-plot)   

```{r awdOtpBorno-plot, fig.cap="OTP & AWD cases in Borno state over time"}
awd_monthly <- awd |> 
  mutate(cases=as.numeric(cases)) |> 
  group_by(yr=year(date),mo=month(date)) |> 
  summarise(cases=sum(cases),.groups="drop") |> 
  mutate(
    date= glue("{yr}{mo}") |> ym()
  )


awd_otp_joined <- awd_monthly |> 
  rename(awd_caseload="cases") |> 
  left_join(
    otp_by_state |> 
      filter(State=="Borno") |> 
      rename(otp_caseload = "cases"), by="date"
)

awd_otp_long <- awd_otp_joined |> 
  select(date, State, awd_caseload,otp_caseload) |> 
  pivot_longer(awd_caseload:otp_caseload)

plot_awd_otp_monthly_borno <- awd_otp_long |> 
  ggplot(aes(x=date, y= value, color= name, fill= name, group=name))+
  geom_line()+
    scale_y_continuous(labels = scales::comma)+
  scale_x_date(
    # minor_breaks = "months", 
    breaks= "1 month",
    # major="year",
    labels=scales::date_format("%b%y")
    )+
  labs(x= "date", title= "Borno: New OTP & AWD cases")+
  theme_bw()+
  theme(
    axis.text.x = element_text(angle=90),
    axis.title.x = element_blank()
  )


plot_awd_otp_monthly_borno
```





## Precip & Soil Moisture Over Time

Figure \@ref(fig:tsWaterAWD-plot) displays AWD cases, soil moisture, and precipitation all plotted over time. Note that AWD cases have been divided by 100 for convenience so that 1 y-axis could be fit to the graph.


```{r tsWaterAWD-plot,fig.cap= "Precipitation, Soil Moisture, and AWD cases over time. Dynamic plot, double click to re-set time scale"}




awd_long_mj <- awd %>% 
  filter(ADM2_EN %in% c("Maiduguri","Jere"))

# add year-wk - double checked there is only 1 week collected per year per location
awd_long_mj <- awd_long_mj %>%
  mutate(
    yr_wk=paste0(lubridate::year(date),"_",lubridate::week(date))
  )

awd_long_by_wk <- awd_long_mj %>% 
  mutate(
    date= lubridate::ymd(date),
    cases= as.numeric(cases)
  ) %>% 
  group_by(date) %>% 
  summarise(
    value=sum(cases,na.rm=T),.groups = "drop"
  ) %>% 
  mutate(
    yr_wk=paste0(lubridate::year(date),"_",lubridate::week(date)),
    parameter="awd_cases"
  )
chirps_daily_long <- read_rds(here::here("data/chirps_daily_maidu_jere_diss_gte2016.rds"))
chirps_daily_long <- chirps_daily_long %>% 
  group_by(yr=lubridate::year(date)) %>% 
  mutate(
    cumulative_precip=cumsum(value)
  ) %>% 
  ungroup()
  
smap_chirps_awd_long <- bind_rows(chirps_daily_long %>% ungroup() %>% 
  select(-cumulative_precip,-yr),
  smap,awd_long_by_wk %>% 
    select(-yr_wk)
    )


smap_chirps_awd_wide<- smap_chirps_awd_long %>% pivot_wider(names_from = parameter,values_from = value)

dat_wide_filt <- smap_chirps_awd_wide %>% 
  select(date,precip,ssm,susm,awd_cases) %>% 
  mutate(awd_cases=awd_cases/100) %>% 
  filter(date>="2016-01-01")

dat_xts<-xts(dat_wide_filt,order.by =  dat_wide_filt$date)

precip_awd_dygraph<-dygraph(dat_xts) %>% 
  dySeries("precip", label = "precip", axis=('y'),color="#74add1") %>%
  dySeries("ssm", label = "surface moisture" ,axis=('y'),color="#a6d96a") %>% 
  dySeries("susm", label = "sub-surface moisture", axis=('y'),color="#8c510a")  %>% 
  dySeries("awd_cases", label = "awd_cases", axis=('y'),color="red")  %>% 

  dyRangeSelector(dateWindow = c(min(dat_xts$date),
                                 max(dat_xts$date))) %>% 
  dyOptions(connectSeparatedPoints = TRUE)


precip_awd_dygraph

```




```{r,eval=F}




borno_lga_precip_gte2017 <- monthly_precip_aggs$lga |> 
  filter(ADM2_EN %in% borno_lga_names) |> 
  filter(date>="2017-01-01")


awd_monthly_lga_cases <- awd_long |> 
  group_by(ADM1_EN,ADM2_EN, floor_dt) |> 
  summarise(
    cases= sum(cases, na.rm=T)
  )

awd_monthly_lga_cases |> count()
awd_monthly_lga_cases |> filter(n==0)
awd_monthly_lga_cases |> filter(cases==0)



plot3_awd_precip_lga_facet <- borno_lga_precip_gte2017 |> 
  ggplot(aes(x=date, value))+
  geom_path( color="#94CCDC")+
  geom_path(data=awd_monthly_lga_cases, aes(x=floor_dt, y=cases/10), color="#F49595")+
  scale_x_date(
    # breaks= "1 month",
    # major="year", 
    minor="months",
    # labels=scales::date_format("%b")
    )+
  scale_y_continuous(sec.axis = sec_axis(trans = ~.x*10,name = "AWD Caseload"))+
  facet_wrap(~ADM2_EN)+
  labs(x= "Date",
       y="Precipitation (mm)",
       title = "Monthly LGA AWD caseload & Precipitation (CHIRPS)",
       subtitle = "Borno State")+
  theme_bw()+
  theme(
    axis.text.x = element_text(angle = 90)
  )

plot3_awd_precip_lga_facet

```

## NDVI Over Time

- NDVI exhibits seasonal pattern (Figure \@ref(fig:monthlyNDVIParams-plot), top).
- In addition to extracting NDVI values we have also extracted the monthly difference & Z-Score compared to the 10 year (2000-2010) monthly average.
- There could be arguments for using either Absolute values as well as relative values. The best choice will likely depend on the parameter, the scale, and the details of the research question:
  + **Relative Index Argument**:Strength of adaptive strategies might be more interesting to analyze under the "departure from normal" framework. People who are adapted to where they live - it is only the significant deviation from normal that should affect them
  + **Absolut Index Argument**: There are absolute environmental thresholds. Once thresholds are crossed, the effects can snowball (i.e once the temperature hits x degrees y crop can no longer survive)


```{r monthlyNDVIParams-plot, fig.cap="Monthly NDVI and comparison to monthly 10-year (2000-2010) average (MODIS 16-day composites)" }

ndvi_mj_summarised <- ndvi$lga |> 
  filter(ADM2_EN %in% c("Maiduguri","Jere")) |> 
  group_by(date, category) |> 
  summarise(
    value= mean(value),.groups="drop"
  )
  


plot_modisNDVIParams <- ndvi_mj_summarised |> 
  mutate(
    category=fct_relevel(category, "NDVI","diff_NDVI","z_NDVI"),
    category_label = recode(category,"diff_NDVI"="Difference from 10 year average  (2000-2010)",
                            "z_NDVI"="Z Score derived from 10 year record (2000-2010)",
                            "NDVI"="NDVI"
                               )
  ) |> 
  ggplot(aes(x= date, y= value))+
  geom_line()+
  scale_x_date(
    date_breaks = "1 month", date_labels =  "%m-%Y"
    )+
  geom_hline(yintercept=0, color="red")+
  facet_wrap(~category_label,nrow=3,scales = "free_y")+
  theme(
    axis.text.x = element_text(angle=90)
  )


plot_modisNDVIParams

```



# Relationship Mapping

## Precip + AWD Thresholds

- The below graph (Figure \@ref(fig:precipAWDThreshold-plot)) displays a first attempt to map out environmental thresholds which result in a human impact. In Figure \@ref(fig:precipAWDThreshold-plot) a clear pattern emerges in which the start of the rainy season coincides with the initial and most prominent peak in AWD cases.We can see that this peak  consistently occurs within a tight band of cumulative yearly rainfall, between 127 & 186 mm.
- While the regression/time series modeling development (further down) is still in early phases, I think that this thresholding approach has potential to yield very practical findings and should definitely be explored further after some more model development.

```{r precipAWDThreshold-plot, fig.cap="Yearly cummulative precipitation (light blue, left axis) and AWD cases (orange). There is clear AWD case-load peak aft the start of precipitation each year. This AWD peak consistently appears in a narrow band of cumulative precipitation (between 127 & 186 mm)"}


# chirps_daily_long <- chirps_daily_long %>% 
#   group_by(yr=lubridate::year(date)) %>% 
#   mutate(
#     cumulative_precip=cumsum(value)
#   ) 

yrly_max_case_weeks <- awd_long_by_wk %>% 
  group_by(yr=lubridate::year(date)) %>%
  slice_max(order_by =value, n = 1 ) %>% 
  # filter(value==max(value)) %>% 
  mutate( 
    yrmo = paste0(yr,"_",lubridate::month(date))
  )
awd_first_peaks <- awd_long_by_wk %>% 
  filter(
    date %in%c(yrly_max_case_weeks$date[1:3],
               "2020-07-19",
               "2021-07-11")
  )

plot_threshold_precip_awd <- chirps_daily_long%>% 
  ggplot(aes(x=date,cumulative_precip))+
  geom_line(color="#abd9e9",lwd=1)+
  geom_line(data= awd_long_by_wk,
            aes(x=date,y=value/2),
            lwd=1,color="#fdae61",alpha=0.75)+
 
  labs(y="Yearly cumulative precipitation (mm)")+
  # drawing lines roughly where the rains start each year
  # geom_vline(xintercept=lubridate::ymd("2017-05-01"))+
  # geom_vline(xintercept=lubridate::ymd("2018-04-01"))+
  # geom_vline(xintercept=lubridate::ymd("2019-05-01"))+
  # geom_vline(xintercept=lubridate::ymd("2020-05-01"))+
  # geom_vline(xintercept=lubridate::ymd("2021-05-01"))+
  geom_vline(xintercept=lubridate::ymd("2017-07-09"),color="#f46d43",lwd=1)+
  geom_vline(xintercept=lubridate::ymd("2018-07-15"),color="#f46d43",lwd=1)+
  geom_vline(xintercept=lubridate::ymd("2019-06-30"),color="#f46d43",lwd=1)+
  geom_vline(xintercept=lubridate::ymd("2020-07-19"),color="#f46d43",lwd=1)+
  # geom_vline(xintercept=lubridate::ymd("2020-08-30"),)+
  geom_vline(xintercept=lubridate::ymd("2021-07-11"),color="#f46d43",lwd=1)+
  geom_hline(yintercept=127,color="#4575b4",lwd=1, linetype="dashed")+
  geom_hline(yintercept=184,color="#4575b4",lwd=1, linetype="dashed")+
  # geom_vline(xintercept=lubridate::ymd("2021-08-29"))+
  geom_point(data=awd_first_peaks, aes(x= date, y= value/2),fill="#fdae61", size= 2,pch=21)+
  
  # seems like cases are spiking between 70-300 mm of rain accumulation/
  # first half of the rainy season
  scale_y_continuous(
    sec.axis= sec_axis(trans = ~.x*2,name = "AWD Cases")
  )+
  scale_x_date(
    minor_breaks = "1 month",
    breaks= "1 month",
    # major="year",
    labels=scales::date_format("%b%y")
  )+
  theme_bw()+
  theme(
    axis.text.x = element_text(angle=90,size = 6)
  )+
  annotate(
    "text", label = "127 mm",
    x = lubridate::ymd("2015-10-01"), y = 115, size = 3, colour = "#4575b4"
  )+
  annotate(
    "text", label = "186 mm",
    x = lubridate::ymd("2015-10-01"), y = 200, size = 3, colour = "#4575b4"
  )



plot_threshold_precip_awd
```


## AWD vs OTP
```{r awdVsOtpBorno-plot, fig.cap= "Scatter plot with trendline showing relation between AWD and OTP case-load in Borno State"}
plot_awd_vs_otp <- awd_otp_joined |> 
  ggplot(aes(x=awd_caseload, y=otp_caseload))+
  geom_point(alpha=0.3)+
  labs(x= "AWD Cases", y="New OTP Admissions",title = "AWD vs OTP case-load in Borno State")+
  geom_smooth(stat="smooth",method = "lm",formula = "y~x")+
  theme_bw()


plot_awd_vs_otp


```


## AWD vs precip

- several time scales worth investigating
- Let's start with most granular available -  **weekly** (AWD is weekly)
  + precipitation is aggregated to weekly for comparison (sum total avg rainfall for each week)

```{r}



chirps_weekly <- chirps_daily_long %>% 
  mutate(
    yr= year(date),
    mo= month(date),
    wk= week(date)
  ) %>% 
  group_by(yr,wk,.drop = F) %>% 
  summarise(
    precip = sum(value,na.rm=T),
    num_dry_days= sum(value>=3), .groups = "drop"
  ) %>% 
  pivot_longer(precip:num_dry_days,names_to = "parameter")



awd_weekly_mj_summarised <- awd_long_mj %>% 
  group_by(date= ymd(date) ) |> 
  summarise(
    cases= sum(as.numeric(cases),na.rm=T),.groups="drop"
  ) |> 
  mutate(
    date= ymd(date),
    yr= year(date),
    wk= week(date),
    parameter="awd_cases",
    value= as.numeric(cases)
    
  ) %>% 
  select(date,yr, wk, parameter,value)


awd_chirps_weekly <- bind_rows(chirps_weekly,awd_weekly_mj_summarised)


# this is not quite right for max value of awd since there are some NAs at the end
# awd_chirps_weekly %>% 
#   group_by(parameter) %>% 
#   summarise(
#     min_yr_wk= glue("{min(yr)}-{min(wk)}"),
#     max_yr_wk= glue("{max(yr)}-{max(wk)}")
#   )
  
awd_chirps_gte2017_filt <- awd_chirps_weekly %>% 
  ungroup() %>% 
  filter(yr>=2017) 

awd_chirps_gte2017_filt_wide<- awd_chirps_gte2017_filt %>% 
  pivot_wider(id_cols = c("yr","wk"),names_from=parameter, values_from = value) 
```


- As a first step let's explore the relationship between AWD and precipitation (Figure \@ref(fig:precipAWDWeekly-plot))
  + There is a weak positive relationship evident
  


```{r precipAWDWeekly-plot, fig.cap="Scatter plot with line of best fit - precip vs AWD case-load"}


plot_precip_awd_weekly <- awd_chirps_gte2017_filt_wide |> 
  filter(!is.na(awd_cases)) |> 
  ggplot(aes(x=precip, y=awd_cases))+
  geom_point()+
  geom_smooth(method="lm", formula="y~x")

plot_precip_awd_weekly
```

- Additional investigation shows that the trend-line (particularly the y-intercept) is highly influenced by the year (Figure \@ref(fig:precipAWDWeeklyYr-plot))

```{r precipAWDWeeklyYr-plot, fig.cap="Scatter plot - precip vs AWD caseload colored by year"}

plot_precip_awd_weekly_yrTrends <- awd_chirps_gte2017_filt_wide |> 
  filter(!is.na(awd_cases)) |> 
  ggplot(aes(x=precip, y=awd_cases, color=factor(yr)))+
  scale_color_discrete(name="Year")+
  geom_point()+
  geom_smooth(method = "lm", formula="y~x")


plot_precip_awd_weekly_yrTrends
```

- The strong apparent relationships warrant further investigation. 
- Lets look into regression options
  <!-- + linear regression assumes data are distributed normally - that ain't normal -->



```{r, eval=F}



chirps_weekly %>% 
  filter(parameter=="precip") |> 
  filter(value>0) |> 
  ggplot(aes(x= value))+
  geom_histogram(bins = 50)+
  # scale_x_sqrt(labels=scales::label_number())+
  scale_x_log10(labels=scales::label_number())+
  # scale_y_log10(labels=scales::label_number())+
  labs(x="Histogram of daily rainfall values")


monthly_precip_mj <- monthly_precip_aggs$lga |> 
  filter(ADM2_EN %in% c("Borno","Jere")) |> 
  group_by(date) |> 
  summarise(
    value= mean(value)
  )

monthly_precip_mj |> 
  # filter(value>0) |> 
  filter(date>="2016-01-01") |> 
  ggplot(aes(x=value))+
  geom_histogram()#+
  # scale_x_log10(label=scales::number_bytes)

```


Assumptions:

- Linearity
- Normality of residuals
- homogeneity if residual var
- indep of residual errors

From Figure \@ref(fig:precipAWDWeeklyYr-plot) we can see that the relationship between AWD & precipitation varies from year to year. From Figure \@ref(fig:precipAWDWeeklyYr-plot) we can also see the general trends and approximate fit of each year Therefore, to get started, lets just do a create a simple model for one year. Randomly, I have chosen the year 2020. Below is the model summary as well as the residual model diagnostic plots.

- Here we revisit the 2020 precipitation-AWD trend this time with residuals highlighted (Figure \@ref(fig:precipAWD2020Residuals-plot)).



```{r precipAWD2020Residuals-plot, fig.cap="Precipitation & AWD cases (2020) with trendline and residuals highlighted"}

# make a couple simple models to explore later
m1 <- lm(awd_cases~precip, data=awd_chirps_gte2017_filt_wide)

m2 <- lm(awd_cases~precip, data=awd_chirps_gte2017_filt_wide |> filter(yr==2020))
m3 <- lm(awd_cases~precip+yr, data=awd_chirps_gte2017_filt_wide )
m4 <- lm(log10(awd_cases)~log(precip+1), data=awd_chirps_gte2017_filt_wide |> filter(yr==2020))
m2a <- broom::augment(m2)



plot_awd_precip_2020_w_residuals <- 
  m2a |> 
  ggplot(aes(x=precip,y=awd_cases))+
  geom_point()+
  stat_smooth(method="lm",se=F, formula="y~x")+
  geom_segment(aes(xend=precip,yend=.fitted),color="red", size= 0.3)



plot_awd_precip_2020_w_residuals 

```

- Print out of linear model summary (Table: \@ref(tab:m2Summary-table))

```{r m2Summary-table}
m2_summary_table <- summary(m2) |> 
  gtsummary::tbl_regression() |>
  # gtsummary::add_glance_table() |>
  gtsummary::add_glance_source_note() |> 
  gtsummary::modify_caption(caption = "Model summary")
  


m2_summary_table

```

- Further, we can look at some model diagnostic plots (Figure\@ref(fig:m2ResidDiags-plot))
- It is clear that there is a relationship embedded inte data. However, the simple linear regression as constructed is violating some linear regression rules/assumptions. Therefore these estimates are not reliable. There are several steps we can try to 
  + data transformations - data sets should be normally distributed. Can try some transforms to better approach normal.
  + changing temporal aggregation level - this can change distribution of data in univariate and bivariate space which will also change the residual error terms.
  + adding variables to MLR - this can change multivariate space and re-distribute residual errors. In doing so auto correlation issues can be fixed.
  + ARIMA & GLM techniques to account for auto-correlation through time.

```{r m2ResidDiags-plot, fig.cap="Residual diagnostic plots - Precipitation vs AWD (2020)"}

par(mfrow=c(2,2))
plot(m2 )

```


## AWD vs Soil Moisture

- Explore relationship here

```{r}

precip_sm_awd_wide_gte2017<- smap_chirps_awd_long %>% 
  mutate(
    yrmo= ym(glue("{year(date)}-{month(date)}"))
  ) %>% 
  group_by(yrmo,parameter) %>%
  summarise(
    mean_val=mean(value,na.rm=T),
    sum_val=max(value,na.rm=T),.groups="drop"
  ) %>% 
  mutate(value=if_else(parameter %in%c("awd_cases","precip"),sum_val, mean_val)) %>% 
  select(yrmo, parameter, value) %>% 
  pivot_wider(names_from = parameter,values_from = value) %>% 
  filter(yrmo>="2017-01-01")
precip_sm_awd_wide_gte2017 %>% 
  group_by(yr=year(yrmo)) %>% 
  mutate(cumulative_precip = cumsum(precip),
         deriv1= cumulative_precip-lag(cumulative_precip),
         deriv2= deriv1-lag(deriv1),
         col_deriv2 = if_else(is.na(deriv2)|deriv2<0,"red","green"),
         # deriv3 = c(F,F,diff(diff(cumulative_precip)>0)!=0)
         cum_cat= case_when(
           cumulative_precip
         )
         ) %>% 
  ungroup() %>% 
  ggplot(aes(x=yrmo, y= cumulative_precip, color=col_deriv2, group=cumulative_precip))+
  geom_line(lwd=2)#+
  scale_color_continuous("rdBl")

  
  
precip_sm_awd_wide_gte2017 <- precip_sm_awd_wide_gte2017 %>% 
  group_by(yr=year(yrmo)) %>% 
  mutate(cumulative_precip = cumsum(precip),
         cum_cat= case_when(
           cumulative_precip< 120 ~"too little",
           cumulative_precip< 190 ~"Goldilocks",
           cumulative_precip> 190 ~"too much",
           
         ),
         lag1 = lag(precip,n = 1),
         lag2 = lag(precip,n = 2)
         ) |> 
  ungroup()
  
#############################################
# Detrending
#############################################

# by differencing
precip_sm_awd_wide_gte2017 <- precip_sm_awd_wide_gte2017 |> 
  mutate(
    awd_cases_detrend = awd_cases-lag(awd_cases),
    precip_detrend = precip-lag(precip)
         ) 
# by model fitting

awd_trend_model<-lm(awd_cases~yrmo,precip_sm_awd_wide_gte2017)
awd_trend_model_preds <- broom::augment(awd_trend_model) 

awd_trend_model_preds |> 
  # mutate(awd_detrended_from_model= awd_cases-.fitted) |> 
  ggplot(aes(x=yrmo,y=.fitted))+
  geom_line()+
  geom_line(data=awd_trend_model, aes(x=yrmo, y=awd_cases))+
  geom_line(data=awd_trend_model, aes(x=yrmo, y=.resid))

precip_sm_awd_wide_gte2017 <- precip_sm_awd_wide_gte2017 |> 
  left_join(awd_trend_model_preds |> select(yrmo,awd_resids= .resid))

precip_sm_awd_wide_gte2017|> 
  ggplot(aes(x=yrmo,y=awd_cases_detrend))+
  geom_line(color="green",lwd=1.2)+
  geom_line(data=precip_sm_awd_wide_gte2017,aes(x= yrmo, y= awd_cases), 
            color="orange",lwd=1.2) +
  theme_bw()

lm(awd_cases~precip+ssm+susm+smp,precip_sm_awd_wide_gte2017) %>% summary()
lm(awd_cases~precip+yrmo,precip_sm_awd_wide_gte2017) %>% summary()
lm(awd_cases~sqrt(precip),precip_sm_awd_wide_gte2017) %>% summary()
lm(awd_cases~sqrt(precip)+yrmo,precip_sm_awd_wide_gte2017) %>% summary()
lm(awd_cases_detrend~precip,precip_sm_awd_wide_gte2017) %>% summary()
lm(awd_cases_detrend~precip_detrend,precip_sm_awd_wide_gte2017) %>% summary()

lm(awd_resids~precip,precip_sm_awd_wide_gte2017) %>% summary()


ck2 <- lm(sqrt(awd_cases)~sqrt(precip)+yrmo,precip_sm_awd_wide_gte2017) 
ck2 %>% summary()
plot(ck2)
car::avPlots(ck2)
lm(awd_cases~precip+cum_cat,precip_sm_awd_wide_gte2017) %>% summary()
lm(awd_cases~precip+cumulative_precip,precip_sm_awd_wide_gte2017) %>% summary()
lm(awd_cases~precip+lag1,precip_sm_awd_wide_gte2017) %>% summary()
lm(awd_cases~precip+lag1+lag2,precip_sm_awd_wide_gte2017) %>% summary()

lm(awd_cases~lag1,precip_sm_awd_wide_gte2017) %>% summary()
lm(awd_cases~lag1+smp,precip_sm_awd_wide_gte2017) %>% summary()
lm(awd_cases~smp,precip_sm_awd_wide_gte2017) %>% summary()
lm(awd_cases~susm,precip_sm_awd_wide_gte2017) %>% summary()
lm(awd_cases~ssm,precip_sm_awd_wide_gte2017) %>% summary()
ck <- lm(awd_cases~sqrt(precip)+ssm,precip_sm_awd_wide_gte2017) 
ck %>% summary()
car::avPlots(ck)


rain_monthly.ts<- ts(precip_sm_awd_wide_gte2017 |> 
                       filter(!is.na(awd_cases)) |> pull(precip),
                     frequency = 12)
awd_cases.ts<- ts(precip_sm_awd_wide_gte2017 |> 
             filter(!is.na(awd_cases)) |> pull(awd_cases),frequency = 12)

astsa::lag2.plot(rain_monthly.ts,awd_cases.ts,52,max.lag = 10,corr=T,smooth=T)
astsa::lag2.plot(awd_cases.ts,rain_monthly.ts,52,max.lag = 10,corr = T,smooth = T)
```

## AWD & ACLED

```{r}
# summarise monthly
acled_mj_summarised <- acled_ne_simp %>% 
  st_drop_geometry() |> 
  filter(ADM2_EN %in% c("Maiduguri","Jere")) |> 
  mutate(
    yrmo= ym(glue("{year(date)}-{month(date)}"))
    ) |> 
  group_by(ADM1_EN,yrmo) |> 
  summarise(
    acled_events= n(), 
    acled_fatalities = sum(fatalities,na.rm=T),.groups="drop"
  )

acled_mj_summarised_long <- acled_mj_summarised |>  
  select(-ADM1_EN) |>   
  pivot_longer(acled_events:acled_fatalities,names_to = "parameter") |> 
  rename(date= "yrmo")

smap_chirps_awd_monthly_long <- smap_chirps_awd_long %>% 
  mutate(
    yrmo= ym(glue("{year(date)}-{month(date)}"))
  ) %>% 
  group_by(yrmo,parameter) %>%
  summarise(
    mean_val=mean(value,na.rm=T),
    sum_val=max(value,na.rm=T),.groups="drop"
  ) %>% 
  mutate(value=if_else(parameter %in%c("awd_cases","precip"),sum_val, mean_val)) %>% 
  select(date=yrmo, parameter, value) 

acled_awd_precip_sm_wide <- bind_rows(smap_chirps_awd_monthly_long,acled_mj_summarised_long) |> 
  pivot_wider(names_from=parameter, values_from = value) |> 
  filter(date>="2017-01-01") 
  
acled_awd_precip_sm_wide |> 
  ggplot(aes(x=acled_events,y=acled_fatalities))+
  geom_point()

acled_awd_precip_sm_wide |> 
  ggplot(aes(x=acled_fatalities,y=awd_cases))+
  geom_point()+
  scale_x_sqrt()

lm(awd_cases~acled_fatalities,acled_awd_precip_sm_wide) |> summary()
lm(awd_cases~acled_fatalities+date,acled_awd_precip_sm_wide) |> summary()
lm(awd_cases~precip,acled_awd_precip_sm_wide) |> summary()
lm(awd_cases~precip+acled_fatalities,acled_awd_precip_sm_wide) |> summary()
lm(awd_cases~acled_events,acled_awd_precip_sm_wide) |> summary()


acled_awd_precip_sm_wide |> 
  ggplot(aes(x=date,y=awd_cases))+
  geom_line(color="orange",lwd=1.2)+
  geom_line(data= acled_awd_precip_sm_wide,aes(x=date,y=acled_fatalities*100),color="red")+
  scale_y_continuous(sec.axis = sec_axis(~.x/100,name = "Fatalities (ACLED)"),name = "AWD Cases")
```

## AWD vs NDVI

## AWD Multiple Linear Regression (MLR)

## Auto-regressive Integrated Moving Average (ARIMA) Model

Here I have just started to explore ARIMA technique. It seems that ARIMA models are typically used for univariate predictions. Therefore, to best understand the utility we need to seek additional details regarding research objectives.

- Below is an example of where an ARIMA model is "auto-fit" (`auto.arima`) for AWD case-load predictions 
  + provide further description of `auto.arima` and 6 
- Should figure out if the predicted values can be constrained to 0 and above

```{r arima1, message=F,warning=F}


# https://otexts.com/fpp2/arima-r.html  - good explanation of auto.arima
# set up ts class object. Weekly data so Freq = 52
awd_weekly_ts <-  ts(awd_weekly_mj_summarised$value,frequency = 52)

# use auto arima to select model with best aic
auto_model_awd_weekly <- forecast::auto.arima(awd_weekly_ts, ic="aic")

#forecast
auto_model_awd_weekly_forecast <-  forecast::forecast(auto_model_awd_weekly)

# plot forecasted data
forecast::autoplot(auto_model_awd_weekly_forecast)+
  labs(x="Time (years)", y="AWD Cases" ,title="Forecasts with ARIMA model (0,1,3)(1,1,0)[52]")+
  theme_bw()


```


# Future Work/ Questions

- sadfa
- asdfa




# Appendix

## Extraction of climatic data {#climateDataExtraction}
Below code is the code which uses using the `rgee` package to extract the data from Google Earth Engine.

```{r}

# will copy code here

```

## Data Cleaning 


### AWD Data Cleaning  {#awdCleaning}

Data cleaning for AWD data set

```{r}
# will copy code here

```

### Malnutrition Data Cleaning {#malnutritionCleaning}

```{r, eval= F}
library(tidyverse)
library(sf)
# compile nutrition sector data from jan 2017 to november 2019 and save to rds.
# no real cleaning done here
# could update once November/December 2021 data is available
# https://www.humanitarianresponse.info/en/operations/nigeria/nutrition/documents/document-type/3w-4w


write_output <- c(T,F)[2]

admins <-  load_nga_admins()
ward <- admins$ward

# wards of interest
woi <- ward %>% 
  filter(ADM2_EN %in% c("Maiduguri","Jere"))


muac_2017_2019 <- readxl::read_excel(here::here("data/NIS_Nutrition_2017_to_2019.xlsx"))

# no site level information
muac_2017_2019 <- muac_2017_2019 |> 
  select(
    State,
    LGA,
    Ward,
    muac_total_screened=MUAC_Screened_TOTAL_Children,
    muac_red = MUAC_Screened_YELLOW,
    new_otp_admissions= OTP_New_Admissions,
    yr= Activity_year,
    mo = `Activity_Month`
    
  )




muac_2021 <- readxl::read_excel(here::here("data/nutrition_sector_5w_jan_-_nov_2021_final.xlsx"),sheet="Jan_Nov 2021", skip=2)
# site level information available so tag it on in case it is useful for ward harmonization
muac_2021 <- muac_2021 |> 
  select(
    State,
    LGA,
    Ward,
    muac_total_screened=`Total Number of Children 6-59 months Screened`,
    muac_red= `RED MUAC Screened`,
    new_otp_admissions= `OTP New Admissions (Total)`,
    yr= `Activity Year`,
    mo = `Activity Month`,
    site_name= `Site Name`,
    site_type = `Site Type`,
    site_id = `Sites ID`,
    lon=`Longitude (E)`,
    lat=`Latitude (N)`
    
  )


# site level information available so tag it on in case it is useful for ward harmonization
muac_2020 <- readxl::read_excel(here::here("data/2020_nutrition_5w_january_-_december_final.xlsm"),
                                sheet = "5Ws", skip=2) |> 
  select(
    State,
    LGA,
    Ward,
    muac_total_screened=MUAC_Screened_TOTAL_Children,
    muac_red= `RED MUAC Screened`,
    new_otp_admissions= `OTP New Admissions`,
    yr= `Activity Year`,
    mo = `Activity Month`,
    site_name= `Site Name`,
    site_type = `Site Type`,
    site_id = `Sites ID`,
    lon=`Longitude (E)`,
    lat=`Latitude (N)`
    
  )


muac <- bind_rows(muac_2021,muac_2020,muac_2017_2019)

if(write_output){
  saveRDS(muac, here::here("data/muac_201701_202111.rds"))
  write_csv(muac, here::here("data/muac_201701_202111.csv"))
}



# to give to NGA office ---------------------------------------------------


muac$LGA %>% unique()
ward_lga_lookup<- admins$ward$ADM2_EN |> 
  set_names(admins$ward$ADM3_EN)

muac_cleaned <- muac %>% 
  mutate(
    # across(.cols= matches("^muac_|^new_"), ~replace_na(.x,0)),
    across(.cols = c("State","LGA","Ward","site_name"),~str_to_title(str_squish(.x))),
    mo= str_to_title(mo),
    date= paste0(yr,"-",mo,"-01"),
    date_dt= lubridate::ymd(date),
    LGA= str_replace_all(LGA, c("Askira-Uba"="Askira/Uba",
                                "Postiskum" ="Potiskum" ,
                                "Gaidam" ="Geidam",
                                "Tarmuwa" ="Tarmua" )),
    # fix these na LGAs that have wards
    LGA = if_else(is.na(LGA),recode(Ward, !!!ward_lga_lookup),LGA)
    
  )
muac_cleaned %>% 
  filter(!is.na(date_dt),!is.na(Ward)|!is.na(site_name)|!is.na(site_id)) %>% 
  filter(is.na(Ward))


woi <- woi %>% 
  unite(col = "admins_concat",c(ADM1_EN,ADM2_EN,ADM3_EN),sep = "_",remove = F)
woi %>% 
  st_drop_geometry() %>% 
  select(matches("ADM\\d_EN")) %>% 
  write_excel_clip()
muac %>% 
  filter(!is.na(site_name))

ward %>% 
  filter(str_detect(ADM3_EN,"musar"))


woi$ADM3_EN %>% sort() %>% write_excel_clip()
woi$admins_concat
wards_partial_clean <- muac_cleaned %>% 
  filter(!is.na(date_dt),!is.na(Ward)) %>% 
  # filter(LGA %in% c("Maiduguri","Jere") & State!="Borno")
  filter(State=="Borno",LGA %in% c("Maiduguri","Jere")) %>%
  unite(col = "admins_concat",State:Ward,sep = "_",remove = F) %>% 
  filter(!admins_concat %in% woi$admins_concat) %>% 
  group_by(State,LGA, Ward, site_name) %>% 
  summarise(
    n=n(),.groups = "drop"
  ) %>% 
  mutate(
    HQ_ward_guess = case_when(
      Ward %in% c("Bolori Ii",
                   "Bolori Ii",
                   "Bolori-Ii",
                   "Bolori 2",
                   "Bolori_ii",
                   "Bolori 2a&2c",
                   "Bolori 2b&D")~"Bolori II",
      
      # apparently there is teachers village in both bolori 1 & 2 - did both
      str_detect(Ward,"^Bolori.*|^Teacher.*")~"Bolori I",
      str_detect(Ward,"^Bulab.+in$|^Bulab.+")~"Bulabulin",
      str_detect(Ward,"^Dala.+")~"Dala",
      str_detect(Ward,"^Dus[um].+|^Muna_eth.*|Farm Centre.*")~"Dusuma",
      str_detect(Ward,"^G[ao]mboru.+|Intersos Clinic Gamboru")~"Gamboru",
      str_detect(Ward,"^Gomari.*|Gangamari|Ngomari")~"Gomari",
      Ward %in% c("Gwange","Gwange I","Gwange 1&2","Gwange-I","Gwnge 1","Gwnge 1")~"Gwange I",
      Ward %in% c("Gwange Ii")~"Gwange II",
      Ward %in% c("Gwange Iii","Gwange 3")~"Gwange III",
      Ward %in% c("Shehuri","Shehuri North","Shehuru North")~"Shehuri North",
      str_detect(Ward,"^Shuwari.*|Shok")~"Shehuri North",
      str_detect(Ward,"^Maisandari.*|Fatima Ali Sheriff")~ "Maisandari" ,
      str_detect(Ward,"^Lam[iu]sula*")~ "Lamisula" ,
      str_detect(Ward,"^Musari*|Maimusari.*")~ "Maimusari" ,
      str_detect(Ward,"^Mash.*mari*|505 Clinic/Hf|Mala Sheriff Camp")~ "Mashamari" ,
      str_detect(Ward,"^202.*|^Mair*")~ "Mairi" ,
      
      #,https://reliefweb.int/sites/reliefweb.int/files/resources/referral_pathway_mmc_jere_final_24.03.17.pdf
      str_detect(Ward,"^Gw[oa]zar*")~ "Dusuma" ,
      str_detect(Ward,"^Sabon.*|^Bale.*|Galtimari|^Gs.+imari$|^Jiddari.*")~ "Galtimari",
      str_detect(Ward,"^Dalori.*")~"WRONG LGA",
      str_detect(Ward,"^Fariah.*lock*|^Madinatu.*")~"Old Maiduguri",
      str_detect(Ward,"^K.+mari$")~"Khadammari",
      str_detect(Ward,"^Gajigana*")~"WRONG LGA",
      
      # Gajigana  - there is a ward called Gajigana in Magumeri... maybe it should b
      TRUE~ Ward
    )

  ) %>% 
  unite(col = "admins_concat_new",c(State,LGA,HQ_ward_guess),sep = "_",remove = F) %>% 
  mutate(
    HQ_ward_guess = case_when(
      admins_concat_new =="Borno_Jere_Gamboru"~"WRONG LGA",
      admins_concat_new =="Borno_Jere_Gwange I"~"WRONG LGA",
      admins_concat_new =="Borno_Jere_Maisandari"~"WRONG LGA",
      admins_concat_new =="Borno_Maiduguri_Gomari"~"WRONG LGA",
      admins_concat_new =="Borno_Maiduguri_Dusuma"~"WRONG LGA",
      TRUE~HQ_ward_guess
      ),
    HQ_LGA_guess = case_when(
      str_detect(Ward,"^Dalori.*")~"Konduga",
      admins_concat_new =="Borno_Jere_Gamboru"~"Maiduguri",
      admins_concat_new =="Borno_Jere_Gwange I"~"Maiduguri",
      admins_concat_new =="Borno_Jere_Maisandari"~"Maiduguri",
      admins_concat_new =="Borno_Maiduguri_Old Maiduguri"~"Jere",
      admins_concat_new= str_detect(Ward,"^Gajigana*")~"Magumeri",
      admins_concat_new =="Borno_Maiduguri_Gomari"~"Jere",
      admins_concat_new =="Borno_Maiduguri_Dusuma"~"Jere",
      TRUE ~ NA_character_
    )
    
  ) %>% 
  mutate(matching= admins_concat_new %in% woi$admins_concat) %>% 
  # count(HQ_ward_guess) %>% 
  print(
    n=306
  )


wards_partial_clean %>% 
  mutate(
    HQ_ward_guess2= case_when(
      matching|HQ_ward_guess=="WRONG LGA"|!is.na(HQ_LGA_guess)~HQ_ward_guess,
      TRUE~NA_character_
    )
  ) %>% 
  select(State,LGA,Ward,site_name,HQ_ward_guess=HQ_ward_guess2,HQ_LGA_guess) #%>% 
  filter(is.na(HQ_ward_guess)) %>% 
  count(Ward)
  
  
ward_harm <- readxl::read_excel(here::here("data/20220121_muac_maidu_jere_ward_harmonizer_from_nga_team.xlsx"),"name_harmonizer")

ward_dictionary <- ward_harm %>%
  filter(!is.na(ward_correction),(is.na(LGA_correction)|LGA_correction %in% c("Maiduguri","Jere"))) %>% 
  mutate(
    LGA_crct= if_else(is.na(LGA_correction),LGA,LGA_correction)
  ) %>% 
  unite(col ="admin_key", c(State,LGA,Ward),sep = "_",) %>% 
  select(admin_key,LGA_crct,ward_crct=ward_correction)

ward_key <- ward_dictionary$ward_crct %>% 
  set_names(ward_dictionary$admin_key)

lga_key <- ward_dictionary$LGA_crct %>% 
  set_names(ward_dictionary$admin_key)

muac_cleaned_wards_harmonized <- muac_cleaned %>% 
  unite(col ="admin_key", c(State,LGA,Ward),sep = "_",remove = F) %>% 
  mutate(
    ward_crct = if_else(admin_key %in% names(ward_key),recode(.x = `admin_key`,!!!ward_key),Ward),
    LGA_crct = if_else(admin_key %in% names(lga_key),recode(.x = `admin_key`,!!!lga_key),LGA)
  )

write_rds(muac_cleaned_wards_harmonized,here::here("data/20220121_MUAC_w_maidu_jere_wards_harmonized.rds"))




```
