---
title: "NE Nigeria Exploratory"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r libraries}
library(tidyverse)
library(rgee)
library(sf)
invisible(purrr::map(list.files(here::here("R"),full.names = T),.f = source))

```


##Load data

```{r load_data}

ee_Initialize()

# 16 day ndvi composites
ndvi_dat_src <- "MODIS/006/MOD13Q1"
ndvi <- ee$ImageCollection(ndvi_dat_src)

# ee_get_date_ic(ndvi)

gee_dat_src <- "UCSB-CHG/CHIRPS/DAILY"
chirps_src <- "UCSB-CHG/CHIRPS/DAILY"
gee_dat <- ee$ImageCollection(gee_dat_src)
chirps <- ee$ImageCollection(chirps_src)

awd <- readxl::read_excel(here::here("data/Borno_EWARS_awd_under5_2017_2021.xlsx")) |> 
  rename(ADM2_EN = `...1`)

muac <- read_rds(file = here::here("data/muac_201701_202111.rds"))
admins <- load_nga_admins()
```

## Clean up data
### AWD data

- AWD data is just caseloads from Borno state. 
- Should make LGA names match shapefile
```{r}

ne_lgas<- admins$lga |> 
  st_drop_geometry() |> 
  filter(ADM1_EN %in% c("Yobe","Borno","Adamawa")) 

awd <- awd |> 
  mutate(
    ADM2_EN= str_replace_all(string = ADM2_EN,c("Askira Uba"="Askira/Uba","Kala Balge"="Kala/Balge" ))
  )
awd_long <- awd |> 
  pivot_longer(-1,names_to = "date",values_to = "cases") |> 
  left_join(ne_lgas |> select(ADM1_EN,ADM2_EN)) |> 
  select(ADM1_EN,ADM2_EN,everything())

```

### MUAC data
- remove records with no muac screenings
- create date time
```{r}
muac |> 
  filter(is.na(new_otp_admissions)& !is.na(muac_total_screened))

muac_filt <- muac |> 
  filter(
    !is.na(new_otp_admissions)|(is.na(new_otp_admissions) & !is.na(muac_total_screened)),
    !is.na(State) 
    ) 
# there are some records with LGA missing, but ward present- should recode
ward_lga_lookup<- admins$ward$ADM2_EN |> 
  set_names(admins$ward$ADM3_EN)

muac_cleaned <- muac_filt|> 
  mutate(
    across(.cols= matches("^muac_|^new_"), ~replace_na(.x,0)),
    across(.cols = c("State","LGA","Ward"),~str_to_title(.x)),
    mo= str_to_title(mo),
    date= paste0(yr,"-",mo,"-01"),
    date_dt= lubridate::ymd(date),
    pct_muac_red= (muac_red/muac_total_screened)*100,
    LGA= str_replace_all(LGA, c("Askira-Uba"="Askira/Uba",
                                "Postiskum" ="Potiskum" ,
                                "Gaidam" ="Geidam",
                                "Tarmuwa" ="Tarmua" )),
    # fix these na LGAs that have wards
    LGA = if_else(is.na(LGA),recode(Ward, !!!ward_lga_lookup),LGA)
    
    )

```
### Admin data
- disregard wards for now
- simplify file. Complex polys are difficult for GEE, especially when stored in memory (can upload as asset later)
```{r}
admins_simp <- admins |> 
  list_modify("ward"=NULL) |> 
  map(~.x |> 
        filter(ADM1_EN %in% c("Borno","Adamawa","Yobe")) |> 
        st_simplify(dTolerance = 0.7)
  )

admins_simp$overall <- admins_simp$state |> 
  summarise() |> 
  mutate(ADM= "overall")# little hack for later
```

## Health Data overview
### Muac overview

- case breakdown
```{r}
muac_long1 <- muac_cleaned |> 
  select(-yr, -mo, -date,-pct_muac_red ) |> 
  pivot_longer(c("muac_total_screened","muac_red","new_otp_admissions") )


plot_muac_cases_overall <- muac_long1|> 
  group_by(name,date_dt,.drop=F ) |> 
  summarise(
    cases=sum(replace_na(value,0))
  ) |> 
  ggplot(aes(x=date_dt,y=cases, fill=name, color=name, group= name))+
  geom_line()+
    scale_y_continuous(labels = scales::comma)+
  scale_x_date(
    # minor_breaks = "months", 
    breaks= "1 month",
    # major="year",
    labels=scales::date_format("%b%y")
    )+
  labs(x= "date", title= "MUAC data -Overall Caseload NE NGA")+
  theme_bw()+
  theme(
    axis.text.x = element_text(angle=90)
  )
plot_muac_cases_overall

```

- # of cases screened for MUAC classified as red
```{r}
muac_long1|> 
  group_by(State,name,date_dt,.drop=F ) |> 
  summarise(
    cases=sum(replace_na(value,0))
  ) |> 
  filter(name=="muac_red") |> 
  ggplot(aes(x=date_dt,y=cases, fill=State, color=State, group= State))+
  geom_line()+
  scale_y_continuous(labels = scales::comma)+
  scale_color_manual(values = c("#959596","#F49595","#BDB7A6"))+
  scale_x_date(
    # minor_breaks = "months", 
    breaks= "1 month",
    # major="year",
    labels=scales::date_format("%b%y")
    )+
  labs(x= "date", title= "# Muac Screenings Classified as Red by State")+
  theme_bw()+
  theme(
    axis.text.x = element_text(angle=90)
  )
```

- % MUAC screened classified as red
```{r}
muac_long2 <- muac_cleaned |> 
  select(-yr, -mo, -date ) |> 
  pivot_longer(c("muac_total_screened","muac_red","new_otp_admissions","pct_muac_red") )

muac_long2 |> 
  filter(name=="pct_muac_red") |> 
    group_by(State,name,date_dt,.drop=F ) |> 
  summarise(
    cases=mean(replace_na(value,0))
  ) |> 
  ggplot(aes(x=date_dt,y=cases, fill=State, color=State, group= State))+
  geom_line()+
  scale_y_continuous(labels = scales::comma)+
  scale_color_manual(values = c("#959596","#F49595","#BDB7A6"))+
  scale_x_date(
    # minor_breaks = "months", 
    breaks= "1 month",
    # major="year",
    labels=scales::date_format("%b%y")
    )+
  labs(x= "date", title= "% MUAC Screenings Classified as Red by State")+
  theme_bw()+
  theme(
    axis.text.x = element_text(angle=90)
  )
```

- **New OTP admissions** - Indicator of choice from this data set
```{r}
muac_long2 |> 
  filter(name=="new_otp_admissions") |> 
    group_by(State,name,date_dt,.drop=F ) |> 
  summarise(
    cases=sum(replace_na(value,0))
  ) |> 
  ggplot(aes(x=date_dt,y=cases, fill=State, color=State, group= State))+
  geom_line()+
  scale_y_continuous(labels = scales::comma)+
  scale_color_manual(values = c("#959596","#F49595","#BDB7A6"))+
  scale_x_date(
    # minor_breaks = "months", 
    breaks= "1 month",
    # major="year",
    labels=scales::date_format("%b%y")
    )+
  labs(x= "date", title= "# New OTP admissions by State")+
  theme_bw()+
  theme(
    axis.text.x = element_text(angle=90)
  )
```

- New OTP admissions in Borno State (by LGA)
```{r}


muac_long2 |> 
  filter(name=="new_otp_admissions") |> 
  filter(State=="Borno") |> 
  group_by(LGA) |> 
  mutate(
    lga_total= sum(replace_na(value,0))
  ) |> 
  group_by(State) |> 
  mutate(
    state_total= sum(replace_na(value,0)),
    lga_pct= lga_total/state_total
    ) |>
  filter(lga_pct>0.03) |> 
  ungroup() |> 
  group_by(LGA,name,date_dt,.drop=F ) |> 
  summarise(
    cases=sum(replace_na(value,0))
  ) |> 
  ggplot(aes(x=date_dt,y=cases, fill=LGA, color=LGA, group= LGA))+
  geom_line()+
  scale_y_continuous(labels = scales::comma)+
  scale_x_date(
    # minor_breaks = "months", 
    breaks= "1 month",
    # major="year",
    labels=scales::date_format("%b%y")
    )+
  labs(x= "date", title= "Borno: New OTP admissions by LGA",subtitle = "LGAs with > 3 % Total Borno OTP admissions")+
  theme_bw()+
  theme(
    axis.text.x = element_text(angle=90)
  )




```
### AWD overview
```{r}
# clean up some classes & round months for monthly agg of AWD cases
awd_long <- awd_long |> 
  mutate(
    cases= as.numeric(cases),
    date_dt= lubridate::ymd(date),
    floor_dt= lubridate::floor_date(date_dt, "month")
  )

# AWD cases per month
awd_monthly_long <- awd_long |> 
  group_by(floor_dt) |> 
  summarise(
    cases=sum(cases,na.rm=T)
  ) 
```


### AWD vs MUAC
```{r}

new_otp_borno_monthly <- muac_long2 |> 
  filter(name=="new_otp_admissions") |> 
  filter(State=="Borno") |> 
  group_by(date_dt) |> 
  summarise(
    new_otp_admissions= sum(replace_na(value,0))
  ) |> 
  rename(floor_dt="date_dt")

awd_otp_joined <- awd_monthly_long |> 
  rename(awd_cases= "cases") |> 
  left_join(new_otp_borno_monthly)

awd_otp_long <- awd_otp_joined |> 
  pivot_longer(-floor_dt)

windows();awd_otp_long |> 
  ggplot(aes(x=floor_dt, y= value, color= name, fill= name, group=name))+
  geom_line()+
    scale_y_continuous(labels = scales::comma)+
  scale_x_date(
    # minor_breaks = "months", 
    breaks= "1 month",
    # major="year",
    labels=scales::date_format("%b%y")
    )+
  labs(x= "date", title= "Borno: New OTP & AWD cases")+
  theme_bw()+
  theme(
    axis.text.x = element_text(angle=90)
  )

awd_otp_joined |> 
  ggplot(aes(x=awd_cases, y=new_otp_admissions))+
  geom_point(alpha=0.3)+
  labs(x= "AWD Cases", y="New OTP Admissions")+
  theme_bw()

```

# GEE Aggregation
### CHIRPS
- Retrieve daily CHIRPS data from 1981 to 2021
- For each month (1981-2021) calculate the sum of the precipitation to get monthly precip
```{r, eval=F}

monthly_precip <- yr_mo_composite_stat_ic(dat = chirps,
                                          month_range = c(1,12),
                                          year_range = c(1981,2021),
                                          stat = "sum")

monthly_precip |> ee_get_date_ic()

monthly_precip |> ee_print()

monthly_precip$first()$getInfo()
ndvi_monthly_2010_2015$first()$getInfo()

monthly_precip_renamed <- monthly_precip$ 
  # CHIRPS release is delayed by a month or 2 so you need to pre-filter
  # or face difficulty later
  filterDate("1981-01-01","2021-11-30") |> 
  ee$ImageCollection$map(
    function(x) {
      date <- ee$Date(x$get("system:time_start"))$format('YYYY_MM_dd')
      name <- ee$String$cat("pr_", date)
      x$select("precipitation")$rename(name)
    })


```

- extract median monthly precipitation (mm) at admin 1 , admin 2, and overall (Borno + Yobe + Adamawa)
- clean up output to tidy long format
- save `.rds` file so expensive calcs can be skipped in future.

```{r, eval=F}
keep_cols <-  c("ADM1_EN","ADM2_EN","ADM")
system.time( # why taking so long... 348 s? 
mo_precip_agg <- admins_simp |> 
  map2( .y = keep_cols,
    ~ ee_extract(
    x = monthly_precip_renamed,
    y = .x[.y],
    scale = 5500,
    fun = ee$Reducer$median(),
    sf = TRUE
  )
)
)

mo_precip_agg_long <- mo_precip_agg |> 
  map(
    ~.x |> 
       st_drop_geometry() |> 
      pivot_longer(-1) |> 
      mutate(
        date= str_remove(string = name, pattern = ".*pr_") |> 
          str_replace_all("_","-") |> lubridate::ymd()

  ) 
  )
# mo_precip_agg_long |>
#   saveRDS(file = here::here("data/monthly_precip_aggs.rds"))


```



```{r}
monthly_precip_aggs <- read_rds(here::here("data/monthly_precip_aggs.rds"))


# left over from checking that no yr_mo_composite_stat function gave same results - all good

# ck1 <- monthly_precip_aggs$overall |> 
#   left_join(
#     mo_precip_agg_long$overall |> 
#       rename(value_new= "value")
#   )
# all(ck1$value==ck1$value_new)
```

- **Malnutrition factors** 
  + might not be directly related to precipitation
  + could have a longer lag time?
  + perhaps more related to famine conditions
    + consecutive dry days/# dry days in month (need to figure out most sensible way to spatial aggregate)
    + Land surface temperature - should pull some references, but I think this has been shown to have predictive power related to agricultural decline
  + z-scores/comparison to historical could be important as they could be influenced by adaptation strategies.

### PDSI (TERRA)
```{r}




terra <- ee$ImageCollection("IDAHO_EPSCOR/TERRACLIMATE")

terra_temp_filt <- terra$
  filterDate("2000-01-01","2020-12-01") |> 
  ee$ImageCollection$map(
    function(x) {
      date <- ee$Date(x$get("system:time_start"))$format('YYYY_MM_dd')
      name <- ee$String$cat("pdsi_", date)
      x$select("pdsi")$rename(name)
    })

```


- extract to admins

```{r}
cat("terra pdsi resolution:", terra_temp_filt$first()$projection()$nominalScale()$getInfo(),"m")


system.time( # ~80 s on 8 gb machine
mo_pdsi_agg <- admins_simp |> 
  map2( .y = keep_cols,
    ~ ee_extract(
    x = terra_temp_filt,
    y = .x[.y],
    scale = 4638,
    fun = ee$Reducer$median(),
    sf = TRUE
  )
)
)


mo_pdsi_long <- mo_pdsi_agg |> 
  map(
    ~.x |> 
       st_drop_geometry() |> 
      pivot_longer(-1) |> 
      mutate(
        date= str_remove(string = name, pattern = ".*pdsi_") |> 
          str_replace_all("_","-") |> lubridate::ymd()

  ) 
  )


mo_pdsi_long$overall |> 
  ggplot(aes(x=date,y =value))+
  geom_line()
```

### NDVI (MODIS 16 day)
- MODIS 250 m 16 day composites 2000-2021.
- 2 images/mo
  + monthly mean composites (mean of 2 images at pixel level)
  + for period of interest (approximately 2017-2021) subtract **10 year** average from monthly composite?
  

```{r}


ndvi |> ee_get_date_ic()

ndvi_only <-  ndvi$select("NDVI")
ndvi_only |> ee_print()
ndvi_only|> ee_get_date_ic() |> pull(time_start) |> range()
ndvi_only$first()$getInfo()



month_sequence <- ee$List$sequence(1,12)
year_sequence <-  ee$List$sequence(2016, 2021)
ndvi_10_yr_mo <- month_sequence$map(
   ee_utils_pyfunc(
     function(m) {
       ndvi_mean <- ndvi_only$filter(ee$Filter$calendarRange(2000, 2010, 'year'))$
         filter(ee$Filter$calendarRange(m, m, 'month'))$
         mean()$
         rename("NDVI_mean")$
         set("system:time_start", ee$Date$millis(ee$Date$fromYMD(2000,m,1)) )
       
       ndvi_std <- ndvi_only$filter(ee$Filter$calendarRange(2000, 2010, 'year'))$
         filter(ee$Filter$calendarRange(m, m, 'month'))$
         reduce(ee$Reducer$stdDev())$
         rename("NDVI_std")$
         set("system:time_start", ee$Date$millis(ee$Date$fromYMD(2000,m,1)) )
       
       
       ndvi_mean$addBands(ndvi_std)
     }

)
)

ndvi_10_yr_mo <- ee$ImageCollection$fromImages(ndvi_10_yr_mo$flatten()) 

ndvi_10_yr_mo |> ee_print()
ndvi_mo_2016_2021<- yr_mo_composite_stat_ic(dat = ndvi_only,
                                                 month_range = c(1,12),
                                                 year_range = c(2016,2021),stat = "mean")
ndvi_mo_2016_2021 |> ee_print()
ndvi_mo_2016_2021
# so what we want to do... 
# for each composite 2016-2020 we want to subtrack the monthly 10 year composite... thats what we want to do.
recent_sub_hist_list <- year_sequence$map(
  ee_utils_pyfunc(
     function(y) {
       month_sequence$map(
         ee_utils_pyfunc(
           function(m){
             recent_yrmo<- ee$Image(ndvi_mo_2016_2021$
              filter(ee$Filter$calendarRange(y, y, 'year'))$
              filter(ee$Filter$calendarRange(m, m, 'month'))$
                first()
             )
             hist_mo <- ee$Image(ndvi_10_yr_mo$
                                   filter(ee$Filter$calendarRange(2000, 2000, 'year'))$
                                   filter(ee$Filter$calendarRange(m, m, 'month'))$
                                   first()
               )
             
            
            NDVI_diff <- recent_yrmo$select("NDVI")$
              subtract(hist_mo$select("NDVI_mean"))$
              rename("NDVI_diff")
            
            NDVI_z <- NDVI_diff$select("NDVI_diff")$
              divide(hist_mo$select("NDVI_std"))$
              rename("NDVI_Z")
            
            recent_yrmo$addBands(NDVI_diff)$addBands(NDVI_z)$
              set("system:time_start",ee$Date$millis(ee$Date$fromYMD(y,m,1)))
           }
         )
       )
       
     }
  
)
)


recent_sub_hist_list |> ee_print()
recent_sub_hist_ic <- ee$ImageCollection$fromImages(recent_sub_hist_list$flatten())
recent_sub_hist_ic |> ee_print()
recent_sub_hist_ic$size()$getInfo()


recent_sub_hist_ic |> ee_print()
recent_sub_hist_ic$first()$propertyNames()$getInfo()
```

- spatial extraction
```{r}
# interesting that scale changes ... i think it is getting erased and replaced by default values when compositing
cat("MODIS NDVI resolution:", recent_sub_hist_ic$first()$projection()$nominalScale()$getInfo(),"m")
cat("MODIS NDVI resolution:", ndvi_only$first()$projection()$nominalScale()$getInfo(),"m")
cat("MODIS NDVI resolution:", ndvi_mo_2016_2021$first()$projection()$nominalScale()$getInfo(),"m")
cat("MODIS NDVI resolution:", ndvi_10_yr_mo$first()$projection()$nominalScale()$getInfo(),"m")


recent_sub_hist_ic$first()$bandNames()$getInfo()

ndvi_2016_2021_diff_renamed <- recent_sub_hist_ic |> 
  # CHIRPS release is delayed by a month or 2 so you need to pre-filter
  # or face difficulty later
  # filterDate("1981-01-01","2021-11-30") |> 
  ee$ImageCollection$map(
    function(x) {
      date <- ee$Date(x$get("system:time_start"))$format('YYYY_MM_dd')
      # name <- ee$String$cat("pr_", date)
      # x$select("NDVI")$rename(name)
      
      ndvi_bandname <- ee$String$cat("NDVI_", date)
      diff_ndvi_bandname <- ee$String$cat("diff_NDVI_", date)
      z_ndvi_bandname <- ee$String$cat("z_NDVI_", date)
      x$rename(ndvi_bandname,diff_ndvi_bandname,z_ndvi_bandname)
      
    })

ndvi_2016_2021_diff_renamed$first()$bandNames()$getInfo()
ndvi_2016_2021_diff_renamed |> ee_print()

# can i aggregate all 3 band seoarately? for all 3 admins?... i suppose i could double map...
system.time( 
  # approx 85 s for single band  ? 
  # 177.07 s for 3 bands
  # 25 s another time for 3 
ndvi_2016_2021_diff_admin <- admins_simp |> 
  map2( .y = keep_cols, 
    ~ ee_extract(
    x = ndvi_2016_2021_diff_renamed,
    # x = recent_sub_hist_ic,
    y = .x[.y],
    scale = 250,
    fun = ee$Reducer$median(),
    sf = TRUE
  )
)
)

#checkout long term monthly to see what the deal is with standar dev
  
ndvi_10_yr_mo$first()$bandNames()$getInfo()
longterm_monthly_renamed <- ndvi_10_yr_mo |> 

  ee$ImageCollection$map(
    function(x) {
      date <- ee$Date(x$get("system:time_start"))$format('YYYY_MM_dd')
      # name <- ee$String$cat("pr_", date)
      # x$select("NDVI")$rename(name)
      
      mean_ndvi_bandname <- ee$String$cat("mean_NDVI_", date)
      std_ndvi_bandname <- ee$String$cat("std_NDVI", date)
      
      x$rename(mean_ndvi_bandname,std_ndvi_bandname)
      
    })

system.time( 
  # approx 85 s for single band  ? 
  # 177.07 s for 3 bands
  # 25 s another time for 3 
longterm_monthly <- admins_simp |> 
  map2( .y = keep_cols, 
    ~ ee_extract(
    x = longterm_monthly_renamed,
    # x = recent_sub_hist_ic,
    y = .x[.y],
    scale = 250,
    fun = ee$Reducer$mean(),
    sf = TRUE
  )
)
)

ndvi_2016_2021_diff_admin2 <- admins_simp |> 
  map2( .y = keep_cols, 
    ~ ee_extract(
    x = ndvi_2016_2021_diff_renamed,
    # x = recent_sub_hist_ic,
    y = .x[.y],
    scale = 250,
    fun = ee$Reducer$mean(),
    sf = TRUE
  )
)



ndvi_mo_2016_2021_long <- ndvi_2016_2021_diff_admin |> 
  map(
    ~.x |> 
       st_drop_geometry() |> 
      pivot_longer(-1) |> 
      mutate(
        category= str_extract(string = name,pattern = "diff_NDVI|_NDVI_|z_NDVI") |> 
          str_replace_all(pattern = "_NDVI_","NDVI"),
        date= str_remove(string = name, pattern = ".*NDVI_") |> 
          str_replace_all("_","-") |> lubridate::ymd()
  )
  )

ndvi_mo_2016_2021_long2 <- ndvi_2016_2021_diff_admin2 |> 
  map(
    ~.x |> 
       st_drop_geometry() |> 
      pivot_longer(-1) |> 
      mutate(
       category= str_extract(string = name,pattern = "diff_NDVI|_NDVI_|z_NDVI") |> 
          str_replace_all(pattern = "_NDVI_","NDVI"),
        date= str_remove(string = name, pattern = ".*NDVI_") |> 
          str_replace_all("_","-") |> lubridate::ymd()
  )
  )
long_term_monthly_long <- longterm_monthly |> 
  map(
    ~.x |> 
       st_drop_geometry() |> 
      pivot_longer(-1) |> 
      mutate(
        category= str_extract(string = name,pattern = "mean_NDVI|std_NDVI") ,
        date= str_remove(string = name, pattern = ".*DVI_") |> 
          str_replace_all("_","-") |> lubridate::ymd()
  )
  )

```



```{r}

saveRDS(ndvi_mo_2016_2021_long,
        here::here("data/monthly_ndvi_stats_2016_2021.rds"))

```


```{r}


ndvi_mo_2016_2021_long$overall |> 
  ggplot(aes(x= date, y= value))+
  geom_line()+
  scale_x_date(
    minor_breaks = "1 month",
    breaks= "1 year",
    # major="year",
    labels=scales::date_format("%b")
  )+
  facet_wrap(~category,nrow=3,scales = "free_y")+
  theme(
    axis.text.x = element_text(angle=90)
  )

ndvi_mo_2016_2021_long$overall |> 
  filter(category!= "NDVI") |> 
  ggplot(aes(x= date, y= value))+
  geom_line()+
  scale_x_date(
    minor_breaks = "1 month",
    breaks= "1 year",
    # major="year",
    labels=scales::date_format("%b")
  )+
  geom_hline(yintercept=0, color="red")+
  facet_wrap(~category,nrow=3,scales = "free_y")+
  theme(
    axis.text.x = element_text(angle=90)
  )


windows();ndvi_mo_2016_2021_long$overall |> 
  filter(category=="diff_NDVI") |> 
  ggplot(aes(x= date, y= value))+
  geom_line()+
  scale_x_date(
    minor_breaks = "1 month",
    breaks= "1 year",
    # major="year",
    labels=scales::date_format("%b%y")
  )+
  geom_hline(yintercept=0, color="red")+
  geom_line(data=ndvi_mo_2016_2021_long$overall |> 
  filter(category=="z_NDVI"), aes(x=date,y=value*500),color="orange")+
  scale_y_continuous(
    sec.axis = sec_axis( ~.x/500)
  )

# its odd how Feb looks similar in diff, but different in z

ndvi_mo_2016_2021_long$overall |> 
  filter(category=="diff_NDVI") |> 
  filter(lubridate::month(date)=="3")

ndvi_mo_2016_2021_long$overall |> 
  filter(category=="z_NDVI") |> 
  filter(lubridate::month(date)=="3")
long_term_monthly_long$overall |> 
  # filter(category=="std_NDVI") |> 
  filter(lubridate::month(date)=="3")

# this is annoying i have to find out...
# is it because its the average z score not the average mean - average x/ average sd?
# should experiment with medians vs means?


diff_mar2016 <- 30.9 
diff_mar2017 <- 31.6
z_mar2016 <- 0.463
z_mar2017 <- 0.819
diff_mar2016/x = z_mar2016
sd_supposed <- diff_mar2016/z_mar2016
diff_mar2017/sd_supposed
diff_mar2016/sd_supposed


# its odd how Feb looks similar in diff, but different in z

ndvi_mo_2016_2021_long2$overall |> 
  filter(category=="diff_NDVI") |> 
  filter(lubridate::month(date)=="3")

ndvi_mo_2016_2021_long2$overall |> 
  filter(category=="z_NDVI") |> 
  filter(lubridate::month(date)=="3")

# version with ee$reducer$mean()
windows();ndvi_mo_2016_2021_long2$overall |> 
  filter(category=="diff_NDVI") |> 
  ggplot(aes(x= date, y= value))+
  geom_line()+
  scale_x_date(
    minor_breaks = "1 month",
    breaks= "1 year",
    # major="year",
    labels=scales::date_format("%b%y")
  )+
  geom_hline(yintercept=0, color="red")+
  geom_line(data=ndvi_mo_2016_2021_long$overall |> 
  filter(category=="z_NDVI"), aes(x=date,y=value*500),color="orange")+
  scale_y_continuous(
    sec.axis = sec_axis( ~.x/500)
  )
# 
# ndvi_hist_current_together_long <- bind_rows(
# ndvi_mo_2016_2021_long$overall |> 
#   mutate(
#     category="absolute_value"
#   ),ndvi_diff_long$overall |> 
#   mutate(
#     category="diff from mean"
#   )
# ) 

# windows();ndvi_hist_current_together_long|> 
#   ggplot(aes(x= date, y= value))+
#   geom_line()+
#   scale_x_date(
#     minor_breaks = "1 months",
#     breaks= "1 year",
#     # major="year",
#     labels=scales::date_format("%b")
#   )+
#   facet_wrap(~category,nrow=2,scales = "free_y")+
#   theme(
#     axis.text.x = element_text(angle=90)
#   )

  
set_timestamp <- function(img){
  img$
    set("system:time_start",ee$Date$millis("system:time_start"))
}
```

- scrap trying to figure out using yr_mo_stat function
```{r, eval=F}
# ndvi_tfilt <- ndvi_only$filterDate("2000-02-01","2015-12-31") 
# 
# ndvi_monthly_2000_2015<- yr_mo_composite_stat_ic(dat = ndvi_tfilt,
#                                                  month_range = c(1,12),
#                                                  year_range = c(2000,2015),stat = "mean")

ndvi_monthly_2000_2015<- yr_mo_composite_stat_ic(dat = ndvi_only,
                                                 month_range = c(1,12),
                                                 year_range = c(2000,2015),stat = "mean")

ndvi_monthly_2000_2015 <- ndvi_monthly_2010_2015$
  filterDate("2000-02-01","2015-12-31") 
ndvi_monthly_2000_2015 |> ee_print()


ndvi_monthly_2010_20152$
  filterDate("2000-02-01","2015-12-31") |> 
  ee_print()

ndvi_monthly_2010_2015 |> ee_get_date_ic() 
ndvi_monthly_2010_2015 |> ee_print() 
ndvi_monthly_2010_2015$first()$getInfo()


addTime <- function(image) {
  image$addBands(image$metadata('system:time_start'));
}

ndvi_monthly_2010_2015$map(
       function(image) {
       # image$addBands(ee$Number(image$metadata('system:time_start')));
       image$metadata('system:time_start')$getInfo()
}
)
# month_list$map(
#           ee_utils_pyfunc(function (m) {
ck2 <- ndvi_monthly_2000_2015$map(
   ee_utils_pyfunc(
     function(image) {
       image$set('system:time_start', image$get('system:time_start'))
       # image$addBands((image$metadata('system:time_start')));
       # image$metadata('system:time_start')$getInfo()
}
)
)
ck2$select("system:time_start")$first()$getInfo()

ck3 <- ck2$
  filter(ee$Filter$calendarRange(2000, 2010, 'year'))$
  filter(ee$Filter$calendarRange(1, 1, 'month'))

ck3 <- ndvi_only$
  filter(ee$Filter$calendarRange(2000, 2010, 'year'))$
  filter(ee$Filter$calendarRange(1, 1, 'month'))$
  mean()
ck3$size()$getInfo()
ck3 |> ee_print()
ee$datetime
```



  
# Additional Indicators

- Drought:
  + PSI

\begin{equation}
PSI = {P-Pm\over $\sigma$ }(\#eq:psi)
\end{equation}

Where:

- P = Total precipitation for period (3 month)
- Pm = historical mean for period
- $\sigma$ = Standard deviation for period

also GEE spyder
https://www.un-spider.org/advisory-support/recommended-practices/recommended-practice-drought-monitoring-spi/step-by-step


hey include the Palmer drought severity index (PDSI), surface water supply index (SWSI), reconnaissance drought index (RDI), normalised difference water index (NDWI) and the standardised moisture anomaly index (SMAI) 









### PDSI (TERRA)



#### visual terra
```{r}
st_centroid(admins_simp$overall)
# Define the visualization parameters.
vizParams <- list(
  bands = c("pdsi_2010_01_01"),
  min = -400,
  max = 600,
  palette = c(
  "#d73027", "#f46d43", "#fdae61",
  "#fee08b", "#d9ef8b", "#a6d96a",
  "#66bd63", "#1a9850"
)
  # gamma = c(0.95, 1.1, 1)
)

# terra_temp_filt$first()$bandNames()$getInfo()
# Center the map and display the image.
Map$setCenter(12.47162 ,11.38009, 10) # San Francisco Bay


Map$addLayer(ee$Image(terra_temp_filt$first()), vizParams, "false color composite")

```




### PSI
lets try to calculate 3 month PSI for muac

```{r}
library(tidyquant)
muac_cleaned |> count(State)
monthly_precip_aggs$overall |> 
  arrange(date)


muac_summarised_month <- muac_cleaned |> 
  group_by(date_dt) |> 
  summarise(
    muac_total_screened= sum(muac_total_screened,na.rm=T),
    muac_red= sum(muac_red,na.rm = T)
  ) |> 
  mutate(
    pct_muac_red= muac_red/muac_total_screened
  )
muac_summarised_month |> 
  ggplot(aes(x=date_dt, y=pct_muac_red))+
  geom_path()+
  scale_y_continuous(labels=scales::percent)+
  (minor_breaks = "months")


muac_summarised_month |> 
  ggplot(aes(x=date_dt, y=muac_red))+
  geom_path()+
  scale_y_continuous(labels = scales::comma)+
  scale_x_date(
    # minor_breaks = "months", 
    breaks= "1 month",
    # major="year",
    labels=scales::date_format("%b%y")
    )+
  theme(
    axis.text.x = element_text(angle=90)
  )
```


```{r}

overall_PSI <- monthly_precip_aggs$overall %>%
  filter(date>="2017-01-01") |> 
  tq_mutate(
    select     = value,
    mutate_fun = rollapply,
    width      = 3,
    align      = "right",
    FUN        = sum,
    col_rename = "sum_3_mo"
  ) |> 
  mutate(
    month_fct= lubridate::month(date, label =T)
  ) |> 
  left_join(
    monthly_precip_aggs$overall %>%
  filter(date<"2017-01-01") |> 
    group_by(month_fct= lubridate::month(date, label=T)) |> 
    summarise(
      historical_mean= mean(value,na.rm=T),
      historical_sd = sd(value,na.rm=T)
      
    ) 
  ) |> 
  mutate(
    PSI = (sum_3_mo-historical_mean)/historical_sd
  )

overall_PSI |> 
  ggplot(aes(x=date,y=PSI))+
  geom_path()+
  geom_point()+
  scale_y_continuous()+
  scale_x_date(
    # minor_breaks = "months", 
    breaks= "1 month",
    # major="year",
    labels=scales::date_format("%b%y")
    )+
  theme(
    axis.text.x = element_text(angle=90)
  )
```
# Precip & AWD
## Precip (mm) & AWD Borno State

```{r}
# filter precip to just borno state & records > 2017
borno_gte2017_precip <- monthly_precip_aggs$state |> 
  filter(ADM1_EN =="Borno") |> 
  filter(date>="2017-01-01")

plot_awd_precip <- borno_gte2017_precip |> 
  ggplot(aes(x=date, value))+
  geom_path( color="#94CCDC")+
  geom_path(data=awd_monthly_long, aes(x=floor_dt, y=cases/100), color="#F49595")+
  scale_x_date(
    # breaks= "1 month",
    # major="year", 
    minor="months",
    # labels=scales::date_format("%b")
    )+
  scale_y_continuous(sec.axis = sec_axis(trans = ~.x*100,name = "AWD Caseload"))+
  labs(x= "date",y="Precipitation (mm)", title = "Monthly AWD caseload & Precipitation (CHIRPS)",subtitle = "Borno State")+
  theme_bw()+
  theme(
    axis.text.x = element_text(angle = 90)
  )

plot_awd_precip
```

### Precip & AWD cases Borno by LGA
overview
```{r}
borno_lga_names <- admins$lga |> 
  filter(ADM1_EN %in% "Borno") |> 
  pull(ADM2_EN)

borno_lga_precip_gte2017 <- monthly_precip_aggs$lga |> 
  filter(ADM2_EN %in% borno_lga_names) |> 
  filter(date>="2017-01-01")


awd_monthly_lga_cases <- awd_long |> 
  group_by(ADM1_EN,ADM2_EN, floor_dt) |> 
  summarise(
    cases= sum(cases, na.rm=T)
  )

plot3_awd_precip_lga_facet <- borno_lga_precip_gte2017 |> 
  ggplot(aes(x=date, value))+
  geom_path( color="#94CCDC")+
  geom_path(data=awd_monthly_lga_cases, aes(x=floor_dt, y=cases/10), color="#F49595")+
  scale_x_date(
    # breaks= "1 month",
    # major="year", 
    minor="months",
    # labels=scales::date_format("%b")
    )+
  scale_y_continuous(sec.axis = sec_axis(trans = ~.x*10,name = "AWD Caseload"))+
  facet_wrap(~ADM2_EN)+
  labs(x= "Date",
       y="Precipitation (mm)",
       title = "Monthly LGA AWD caseload & Precipitation (CHIRPS)",
       subtitle = "Borno State")+
  theme_bw()+
  theme(
    axis.text.x = element_text(angle = 90)
  )

plot3_awd_precip_lga_facet

```
- zoom in 

```{r}

lgas_with_awd_cases<-awd_monthly_lga_cases$ADM2_EN |> unique()
plots4_lgas <- lgas_with_awd_cases |> 
  map(
    ~borno_lga_precip_gte2017 |> 
      filter(ADM2_EN==.x) |> 
      ggplot(aes(x=date, value))+
      geom_path( color="#94CCDC")+
      geom_path(data=awd_monthly_lga_cases |> filter(ADM2_EN ==.x), 
                aes(x=floor_dt, y=cases/10), color="#F49595")+
  scale_x_date(
    # breaks= "1 month",
    # major="year", 
    minor="months",
    # labels=scales::date_format("%b")
    )+
  scale_y_continuous(sec.axis = sec_axis(trans = ~.x*10,name = "AWD Caseload"))+
  labs(x= "Date",
       y="Precipitation (mm)",
       title = "Monthly LGA AWD caseload & Precipitation (CHIRPS)",
       subtitle = glue::glue("Borno State - {.x}"))+
  theme_bw()+
  theme(
    axis.text.x = element_text(angle = 90)
  )
  ) |> 
  set_names(lgas_with_awd_cases)


```


```{r}

plots4_lgas$Maiduguri
plots4_lgas$Shani
plots4_lgas$Ngala
plots4_lgas$Monguno
plots4_lgas$Gwoza

```

### Precipitation Z-Scores & AWD Borno
lets try with some precipitation Z scores. No evident pattern - hypothesis: AWD linked to natural phenomena and thresholds. Z-scores might be more appropriate when looking at peoples ability to adapt to certain conditions (i.e flood) where there "normal" is very important. However, microbial/physical processes just need a mechanism to contaminate.

```{r}

monthly_avg_precip_lt_2017 <- monthly_precip_aggs$state |> 
  filter(ADM1_EN =="Borno") |> 
  filter(date<"2017-01-01") |> 
  group_by(month_fct= lubridate::month(date,label = T)) |> 
  summarise(
    mean_precip= mean(value),
    sd_precip= sd(value)
  )

borno_precip_z_gte_2017 <- monthly_precip_aggs$state |> 
  filter(ADM1_EN =="Borno") |> 
  filter(date>="2017-01-01") |> 
  mutate(month_fct=lubridate::month(date, label=T)) |> 
  left_join(monthly_avg_precip_lt_2017) |> 
  mutate(
    precip_z= (value-mean_precip)/sd_precip
  )


awd_precip_z_plot <- borno_precip_z_gte_2017 |> 
  ggplot(aes(x=date, precip_z))+
  geom_path( color="blue")+
  geom_path(data=awd_monthly_long, aes(x=floor_dt, y=cases/1000), color="red")+
  scale_y_continuous(sec.axis = sec_axis(trans = ~.x*1000,name = "AWD Caseload"))+
  labs(x= "date",y="Precipitation (Z score)", title = "Monthly AWD caseload & Precipitation (CHIRPS)",subtitle = "Borno State")+
  theme_bw()

awd_precip_z_plot
```

## Precip & OTP

```{r}

otp_monthly <- muac_long2 |> 
  filter(name=="new_otp_admissions") |> 
    group_by(name,date=date_dt,.drop=F ) |> 
  summarise(
    value=sum(replace_na(value,0)),.groups="drop"
  ) 

precip_monthly_gte2017 <- monthly_precip_aggs$overall |> 
  mutate(
    name= "precipitation (mm)"
  ) |> 
  filter(date>="2017-01-01") 

precip_monthly_gte2017|> 
  bind_rows(otp_monthly) |> 
  ggplot(aes(x=date, y= value))+
  geom_line()+
  facet_wrap(~name,scales = "free_y",nrow = 2)


plot_otp_precip <- precip_monthly_gte2017 |> 
  ggplot(aes(x=date, value))+
  geom_path( color="#94CCDC")+
  geom_path(data=otp_monthly, aes(x=date, y=value/100), color="#F49595")+
  scale_x_date(
    # breaks= "1 month",
    # major="year", 
    minor="months",
    # labels=scales::date_format("%b")
    )+
  scale_y_continuous(sec.axis = sec_axis(trans = ~.x*100,name = "AWD Caseload"))+
  labs(x= "date",y="Precipitation (mm)", title = "Monthly AWD caseload & Precipitation (CHIRPS)",subtitle = "Borno State")+
  theme_bw()+
  theme(
    axis.text.x = element_text(angle = 90)
  )

plot_otp_precip

precip_monthly_gte2017 |> 
  select(-name) |> 
  rename(precip_mm="value") |> 
  left_join(
    otp_monthly |> 
      rename(new_otp="value") |> 
      select(-name),
    by="date") |> 
  ggplot(aes(x=precip_mm,y=new_otp))+
  geom_point()


precip_lga_monthly_gte2017 <- monthly_precip_aggs$lga |> 
  mutate(
    name= "precipitation (mm)"
  ) |> 
  filter(date>="2017-01-01") 


otp_monthly_lga <- muac_long2 |> 
  filter(name=="new_otp_admissions") |> 
    group_by(State,LGA,date=date_dt,.drop=F ) |> 
  summarise(
    value=sum(replace_na(value,0)),.groups="drop"
  ) |> 
  rename(new_otp="value")

precip_otp_lga_monthly_j <- precip_lga_monthly_gte2017 |> 
  rename(
    precip_mm=value
  ) |> 
  left_join(otp_monthly_lga |> rename(ADM2_EN="LGA")) 

precip_otp_lga_monthly_j |> 
  filter(State=="Borno") |> 
  ggplot(aes(x=precip_mm,y=new_otp))+
  geom_point()+ 
  facet_wrap(~ADM2_EN)
```

## Drought (PDSI) & MUAC caseload

```{r}


otp_pdsi_monthly <- mo_pdsi_long$overall |> 
  select(value, date) |> 
  filter(date>"2015-01-01") |> 
  mutate(
    name="PDSI"
  ) |> 
  bind_rows(
    otp_monthly
  )
otp_pdsi_monthly |> 
  ggplot(aes(x=date, y= value,color= name,group=name))+
  geom_line()

otp_pdsi_monthly |> 
  ggplot(aes(x=date, y= value))+
  geom_line()+
  facet_wrap(~name,scales = "free_y",nrow = 2)



  

```



### Drought PDSI Z-Scores
- do historical comparisons at all overall, state, and LGA
```{r}

admin_levels<- list(
     state = c("ADM1_EN","month_fct"),
     lga= c("ADM2_EN","month_fct"),
      overall= "month_fct")


pdsi_historical_comparison <- mo_pdsi_long |> 
  map2(.y = admin_levels,
      ~ .x |> 
          mutate(month_fct= lubridate::month(date,label=T)) |>
         compare_to_historical(
           historical_from = "2000-01-01",
           historical_to = "2010-12-31",
           from= "2015-01-01",
           group_vars = .y)
         
         )
windows();pdsi_historical_comparison$overall |> 
  select(date, pdsi=value, pdsi_diff=value_diff, pdsi_z= value_z) |> 
  pivot_longer(-date) |> 
  bind_rows(
    otp_monthly
  ) |> 
  ggplot(aes(x=date, y= value))+
  geom_line()+
   facet_wrap(~name,
             scales = "free_y",nrow = 4)

pdsi_historical_comparison$overall |> 
  ggplot(aes(x= value, y=value_diff))+geom_point()
pdsi_historical_comparison$overall |> 
  ggplot(aes(x= value, y=value_z))+geom_point()


```

