---
title: "rgee_learning_cheats"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

- scale changes when compositing.
  + don't think it matters as long as you set the scale properly when using subsequent reducers
- `


straight from google
```{r}
# // Make a date filter to get images in this date range.
# dateFilter = ee.Filter.date('2014-01-01', '2014-02-01');

# // Load a MODIS collection with EVI data.
# var mcd43a4 = ee.ImageCollection('MODIS/MCD43A4_006_EVI')
    # .filter(dateFilter);
    
# ndvi with historical bands
summary_stats_ic_moyID <- summary_stats_ic |> 
  ee$ImageCollection$map(
    function(img){
      moy = ee$Date(img$get('system:time_start'))$getRelative('month','year')
      img$set("moy",moy)
    
  
    }
)

# ndvi current bands
ndvi_gte2016_moyID <- ndvi_gte2016 |> 
  ee$ImageCollection$map(
    function(img){
      moy = ee$Date(img$get('system:time_start'))$getRelative('month','year')
      img$set("moy",moy)
    }
)



# // Define an inner join.
innerJoin = ee$Join$inner();

# // Specify an equals filter for image timestamps.
filterMOYEq <- ee$Filter$equals(leftField = 'moy', rightField = 'moy')


# // Apply the join.
innerJoinedNDVI = innerJoin$apply(ndvi_gte2016_moyID, summary_stats_ic_moyID, filterMOYEq)

# // Display the join result: a FeatureCollection.
innerJoinedNDVI |> ee_print()


# // Map a function to merge the results in the output FeatureCollection.
joinedMODIS = innerJoinedNDVI$map(function(feature) {
   ee$Image$cat(feature$get('primary'), feature$get('secondary'));
})

# // Print the result of merging.
joinedMODIS |> ee_print()
joinedMODIS$getInfo()

ee$ImageCollection(joinedMODIS) |> ee_print()
# damn those google guys are 
```


# this might be a stupid wrapper 
```{r}
ic_combine_bands_by<- function(ic1, ic2, by){
 
# // Define an inner join
innerJoin = ee$Join$inner()

# // Specify an equals filter for image timestamps.
filterEq <- ee$Filter$equals(leftField = by, rightField = by)

# // Apply the join.
innerJoined_ic = innerJoin$apply(ic1, ic2, filterEq)


# // Map a function to merge the results in the output FeatureCollection.
# in the JavaScript code-editor this seems to auto-convert/get coerced to ImageCollection
joined_fc = innerJoinedNDVI$map(function(feature)  {
   ee$Image$cat(feature$get('primary'), feature$get('secondary'));
})

# with rgee is seems necessary to explicitly convert
ee$ImageCollection(joined_fc) 
}


```


straing from cesar

```{r}

# exact example
col <- ee$ImageCollection('MODIS/006/MOD13A2')$select('NDVI')
col <- col$map(function(img) {
  doy <- ee$Date(img$get('system:time_start'))$getRelative('day', 'year')
  img$set('doy', doy)
})
distinctDOY <- col$filterDate('2013-01-01', '2014-01-01')

filter1 <- ee$Filter$equals(leftField = 'doy', rightField = 'doy')

join1 <- ee$Join$saveAll('doy_matches')
joinCol <- ee$ImageCollection(join1$apply(distinctDOY, col, filter1))
joinCol |> ee_print()



col <-summary_stats_ic
col <- col$map(function(img) {
  moy <- ee$Date(img$get('system:time_start'))$getRelative('month', 'year')
  img$set('moy', moy)
})
distinctDOY <- ndvi_gte2016

filter1 <- ee$Filter$equals(leftField = 'moy', rightField = 'moy')

join1 <- ee$Join$saveAll('doy_matches')
joinCol <- ee$ImageCollection(join1$apply(distinctDOY, col, filter1))
joinCol |> ee_print()

comp <- joinCol$map(function(img) {
  doyCol = ee$ImageCollection$fromImages(
    img$get('doy_matches')
  )
  # doyCol$reduce(ee$Reducer$median())
  i= ee$Image(doyCol)
  attached_i = ee$Image(img$get("NDVI_mean"))
  i$addBands(attached_i)
  
})
comp |> ee_print()
comp <- joinCol$map(function(img) {
  i = ee$Image(img)
  attached_i = ee$Image(img$get("NDVI_mean"))
  i$addBands(attached_i)
})
comp |> ee_print()






```



going to try the `ee$ImageCollection$combine`
```{r}

# ImageCollection.combin

col <- col$map(function(img) {
  doy <- ee$Date(img$get('system:time_start'))$getRelative('day', 'year')
  img$set('doy', doy)
})
distinctDOY <- col$filterDate('2013-01-01', '2014-01-01')

ndvi_gte2016 |> ee_print()

ndvi_gte2016_moyID <- ndvi_gte2016 |> 
  ee$ImageCollection$map(
    function(img){
      moy = ee$Date(img$get('system:time_start'))$getRelative('month','year')
      indexString = ee$Number(moy)$format('%03d')
      img$set('system:index',indexString)
    }
)
col <- ndvi_gte2016$map(function(img) {
  doy <- ee$Date(img$get('system:time_start'))$getRelative('month', 'year')
  img$set('butterJohn', doy)
})

col$aggregate_array("butterJohn")$getInfo()


summary_stats_ic |> ee_print()
summary_stats_ic_moyID <- summary_stats_ic |> 
  ee$ImageCollection$map(
    function(img){
      moy = ee$Date(img$get('system:time_start'))$getRelative('month','year')
      indexString = ee$Number(moy)$format('%03d')
      img2=img$set('system:index',indexString)
      img2$set("new_prop",moy)
    
  
    }
)

summary_stats_ic$first()$get('system:index')$getInfo()
summary_stats_ic_moyID$first()$get('system:index')$getInfo()
summary_stats_ic$first()$set('system:index',"suckit")$get('system:index')$getInfo()


summary_stats_ic_moyID |> ee_print()
ndvi_gte2016_moyID$aggregate_array("system:index")$getInfo()
ndvi_gte2016_moyID$aggregate_array("new_prop")$getInfo()
summary_stats_ic_moyID$aggregate_array("system:index")$getInfo()
summary_stats_ic$aggregate_array("system:index")$getInfo()

ndvi_gte2016_w_historical_bands<- ndvi_gte2016_moyID$combine(summary_stats_ic_moyID)
ndvi_gte2016_w_historical_bands |> ee_print()

```


```{r cars}
moyList = ee$List(ndvi_10yr_mean$aggregate_array('moy'))


# Map over the list of moys to calculate NDVI differences.
ndviDifFromMeanList = moyList$map(
  ee_utils_pyfunc(
  function(moy){
  #// Get the 20-year mean NDVI image for the given doy.
  ndviMean = ndvi_10yr_mean$filter(ee$Filter$eq('moy', moy))$first()
  #// Get the single-year NDVI image for the given doy.
  ndviYear = ndvi_gte2016$filter(ee$Filter$eq('moy', moy))$first()
  #// Calculate the single-year NDVI difference from 20-year mean NDVI.
  ndviYear$subtract(ndviMean)$
  #  // Set the doy for the image.
    set('moy', moy)
  }
  )
)

ee$ImageCollection$fromImages(ndviDifFromMeanList)|> ee_print()


ee$Image$normalizedDifference

filter1 <- ee$Filter$equals(leftField = 'moy', rightField = 'moy')
join1 <- ee$Join$saveAll('moy_matches')
joinCol <- ee$ImageCollection(join1$apply(ndvi_gte2016, ndvi_10yr_mean, filter1))
# joinCol <- ee$ImageCollection(join1$apply(ndvi_10yr_mean,ndvi_gte2016, filter1))
joinCol |> ee_print()

comp <- joinCol$map(function(img) {
  moyCol = ee$ImageCollection$fromImages(
    img$get('moy_matches')
  )
comp |> ee_print()
  
    moyCol
```

## Some new funcs

- ic_add_month_property
- ic_join_bands

### Old verbose way

```{r}


ndvi <- ndvi$
  select("NDVI","EVI")$
  map(
    ee_utils_pyfunc(
      function(x){x$
          multiply(0.0001)$
          copyProperties(x,x$propertyNames())
          }
    )
  )


ndvi_2000_2010<- ndvi$filterDate("2000-01-01", "2010-12-31")$
  select("NDVI","EVI")

ndvi_gte2016 <- ndvi$filterDate("2016-01-01", "2022-01-01")$select("NDVI","EVI")

 ndvi_10yr_mean<- yr_mo_composite_stat_ic2(dat =ndvi_2000_2010,
                                      month_range = c(1,12),
                                      year_range = c(2000,2010),
                                      stat = "mean",
                                      monthly_stat_per = "range") 
 
 
 ndvi_10yr_sd<- yr_mo_composite_stat_ic2(dat =ndvi_2000_2010,
                                      month_range = c(1,12),
                                      year_range = c(2000,2010),
                                      stat = c("sd"),
                                      monthly_stat_per = "range") 
 ndvi_10yr_min<- yr_mo_composite_stat_ic2(dat =ndvi_2000_2010,
                                      month_range = c(1,12),
                                      year_range = c(2000,2010),
                                      stat = c("min"),
                                      monthly_stat_per = "range") |> 
   ic_add_month_property()
 
 ndvi_10yr_max<- yr_mo_composite_stat_ic2(dat =ndvi_2000_2010,
                                      month_range = c(1,12),
                                      year_range = c(2000,2010),
                                      stat = c("max"),
                                      monthly_stat_per = "range") 

 

month_sequence <- ee$List$sequence(1,12)
year_sequence <- ee$List$sequence(2016,2021)
 summary_stats_ic_l <- month_sequence$map(
  ee_utils_pyfunc(
    function(m){
      #sd
      ndvi_10yr_sd$filter(ee$Filter$calendarRange(m, m, 'month'))$first()$
        addBands(
          # mean
        ndvi_10yr_mean$filter(ee$Filter$calendarRange(m, m, 'month'))$first()
        )$ # min
        addBands(
        ndvi_10yr_min$filter(ee$Filter$calendarRange(m, m, 'month'))$first()
        )$ #max
        addBands(
        ndvi_10yr_max$filter(ee$Filter$calendarRange(m, m, 'month'))$first()
        )
      
    }
    
  ) 
  
 )
 
 
summary_stats_ic <- ee$ImageCollection$fromImages(summary_stats_ic_l) 

# ndvi with historical bands
summary_stats_ic_moyID <- summary_stats_ic |> 
  ee$ImageCollection$map(
    function(img){
      moy = ee$Date(img$get('system:time_start'))$getRelative('month','year')
      img$set("moy",moy)
    
  
    }
)

# ndvi current bands
ndvi_gte2016_moyID <- ndvi_gte2016 |> 
  ee$ImageCollection$map(
    function(img){
      moy = ee$Date(img$get('system:time_start'))$getRelative('month','year')
      img$set("moy",moy)
    }
)

```

### New way with wrappers

```{r}
ndvi <- ndvi$
  select("NDVI","EVI")$
  map(
    ee_utils_pyfunc(
      function(x){x$
          multiply(0.0001)$
          copyProperties(x,x$propertyNames())
          }
    )
  )



ndvi_2000_2010<- ndvi$filterDate("2000-01-01", "2010-12-31")$
  select("NDVI","EVI")
ndvi_gte2016 <- ndvi$filterDate("2016-01-01", "2022-01-01")$select("NDVI","EVI")


# I need to figure out trick to combine reducers so the 4 stats below can be calculated together
 ndvi_10yr_mean<- yr_mo_composite_stat_ic2(dat =ndvi_2000_2010,
                                      month_range = c(1,12),
                                      year_range = c(2000,2010),
                                      stat = "mean",
                                      monthly_stat_per = "range") |> 
   ic_add_month_property()
 
 
 ndvi_10yr_sd<- yr_mo_composite_stat_ic2(dat =ndvi_2000_2010,
                                      month_range = c(1,12),
                                      year_range = c(2000,2010),
                                      stat = c("sd"),
                                      monthly_stat_per = "range") |> 
   ic_add_month_property()
 ndvi_10yr_min<- yr_mo_composite_stat_ic2(dat =ndvi_2000_2010,
                                      month_range = c(1,12),
                                      year_range = c(2000,2010),
                                      stat = c("min"),
                                      monthly_stat_per = "range") |> 
   ic_add_month_property()
 
 ndvi_10yr_max<- yr_mo_composite_stat_ic2(dat =ndvi_2000_2010,
                                      month_range = c(1,12),
                                      year_range = c(2000,2010),
                                      stat = c("max"),
                                      monthly_stat_per = "range") |> 
   ic_add_month_property()

 
 # new func
 ndvi_10yr_stats <- ndvi_10yr_mean |> 
   ic_join_bands(y = ndvi_10yr_sd,by = "moy") |>  
   ic_join_bands(y=ndvi_10yr_max,by = "moy") |> 
   ic_join_bands(y=ndvi_10yr_min,by = "moy") 



ndvi_gte2016_w_historical_bands<- ic_join_bands(x = ndvi_gte2016 |> ic_add_month_property(),
                                                    y = ndvi_10yr_stats |> ic_add_month_property(),
                                                    by = "moy")


# this is still a bit ugly - opened issue wrap simplifier
ndvi_zscore<- ndvi_gte2016_w_historical_bands$map(
  function(img){
    zscore<- img$expression(
      "float((NDVI-NDVI_mean)/(NDVI_stdDev))",
      opt_map= list(NDVI= img$select("NDVI"),
                    NDVI_mean= img$select("NDVI_mean"),
                    NDVI_stdDev= img$select("NDVI_stdDev")
      )
    )$rename("NDVI_z_score")
    img$select("NDVI","EVI")$addBands(zscore)
  }
  
) 

```




### wish list
- subsetting images from ImageCollection using index

```{r pressure, echo=FALSE}
plot(pressure)
```


