---
title: "Sensitivity Analysis of Aggregation Methods"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Intro

How we aggregate climate data makes a difference. How much difference does it make? This may not be a straight forward question. Sometimes it might make sense to aggregate in different ways depending on the region and the indicator of interest.

I have no answer but have copied an example below which may be developed further to bring light to the discussion


# Example 1
Data Sources:
- `cm$lga` Aggregated monthly level rainfall (sum of rainfall over month) aggregated to LGA
- `chirps_daily_long`: Daily rainfall extracted to dissolved Jere & Maiduguri LGAs

If I aggregate these two data sets to be conceptually the same do I get the same values? Not exactly, but pretty close (close enough?).

Is there a way to do sensitivity analysis to better understand impact of aggregation methods

```{r cars}

library(lubridate)
library(glue)
library(dplyr)


chirps_daily_long <- read_rds(here::here("data/chirps_daily_maidu_jere_diss_gte2016.rds"))
cm <- read_rds(here::here("data/monthly_precip_aggs.rds"))


cm$lga %>% 
  filter(ADM2_EN %in% c("Maiduguri","Jere")) %>% 
  group_by(date) %>% 
  summarise(
    value= mean(value,na.rm=T)
  ) %>% 
  filter(date>"2017-01-01")

chirps_daily_long %>% 
  group_by(yr=year(date), mo=month(date)) %>% 
  summarise(
    value= sum(value),.groups = "drop"
  ) %>% 
  mutate(
    date= ymd(glue("{yr}-{mo}-01"))
    
  ) %>% 
  filter(date>"2017-01-01")
```

