---
title: "ARIMA Timeseries"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r libraries}
library(tidyverse)
library(sf)
library(xts)
# devtools::load_all()
invisible(purrr::map(list.files(here::here("R"),full.names = T),.f = source))

```


<!-- https://www.youtube.com/watch?v=uzshNjQbQV8 -->
1.) stationary
  + no trend - constant mean and variance.
  + covariance is independent of time.
  + if we randomly select two periods of time the mean will be the same.
2.) Seasonality
  + Any predictable fluctuation or pattern that repeats over 1-yr period
3.) Autocorrelation
  + vector is correlated with itself (i.e the vector could be correlated with its lag value)
  
ARMA vs ARIMA
extension of ARMA

uses the lag in 1 vector to forecast
Dependent var y is based on multiple lags of same variable (ARMA)

ARIMA has 3 parameters: p,q,d
- **AR** (p) : **Auto Regression**: # of autoregresseive terms
- **I**  (d):  **Differencing**: # of non seasonal differences integrated
- **MA** (q) : **Moving Average** : # of lagged forecast error in eq (MA)

Normal regression: y = a + bX+e
ARIMA is univariate: 

y= a+b(yt-1)+e

in MA we take the lag values of error values


ARMA does not remove satationarity - no differnecing
when using ARIMA we need to convert non-stationary data to stationary




## Stationarity



```{r data}
awd <- load_awd()
chirps_daily_long <- read_rds(here::here("data/chirps_daily_maidu_jere_diss_gte2016.rds"))
# combine maiduguri and jere into one
awd <- awd %>% 
  mutate(
  date= lubridate::ymd(date),
  cases= as.numeric(cases)
  ) %>% 
  group_by(date) %>% 
  summarise(
    value=sum(cases,na.rm=T),.groups = "drop"
  ) %>% 
  mutate(
    yr_wk=paste0(lubridate::year(date),"_",lubridate::week(date)),
    parameter="awd_cases"
  )

awd <- awd %>% 
  mutate(
    value_diff1 = value - lag(value),
    pct_change  = (value- lag(value))/lag(value)
  )


awd_xts <- xts(x = awd[,"value"],order.by = awd$date)
awd2 <- awd %>% filter(!is.na(pct_change))
awd_xts2 <- xts(x = awd2$pct_change,order.by = awd2$date)

chirps_daily_ts <-  xts(chirps_daily_long$value,order.by = chirps_daily_long$date)
```



# Auto-correlation plot
```{r}
library(tseries)
library(lubridate)
library(glue)
library(forecast)
acf(awd_xts)
pacf(awd_xts)
acf(awd_xts2)
pacf(awd_xts2)


```
### Augmented Dickey Fuller Test

```{r}
adf.test(awd_xts)
res2 <- adf.test(awd_xts2)
res2
```

### Auto- ARIMA
Get best model
```{r}

forecast::tsdisplay(awd_xts)
forecast::tsdisplay(chirps_daily_ts)
cm <- read_rds(here::here("data/monthly_precip_aggs.rds"))
cm_monthly <- cm$lga %>% 
  filter(ADM2_EN %in% c("Maiduguri","Jere")) %>% 
  group_by(date) %>% 
  summarise(
    value= mean(value,na.rm=T)
  ) %>% 
  filter(date>="2017-01-01")


cm_monthly$date %>% range()
chirps_monthly_ts <-  ts(cm_monthly$value,frequency = 12)
plot(chirps_daily_ts)

avg_model_chirps_monthly <- forecast::Arima(chirps_monthly_ts,c(0,0,0))
avg_model_chirps_monthly_forecast <-  forecast::forecast(avg_model_chirps_monthly)
plot(avg_model_chirps_monthly_forecast)


auto_model_chirps_monthly <- auto.arima(chirps_monthly_ts)
auto_model_chirps_monthly_forecast <-  forecast::forecast(auto_model_chirps_monthly)
plot(auto_model_chirps_monthly_forecast)


awd_weekly_ts <-  ts(awd$value,frequency = 52)
auto_model_awd_weekly <- auto.arima(awd_weekly_ts, ic="aic")
auto_model_awd_weekly_forecast <-  forecast::forecast(auto_model_awd_weekly)
autoplot(auto_model_awd_weekly_forecast)


awd_weekly_xts <-  xts(awd$value,order.by = awd$date)
auto_model_awd_weekly <- auto.arima(awd_weekly_ts, ic="aic")
auto_model_awd_weekly_forecast <-  forecast::forecast(auto_model_awd_weekly)
autoplot(auto_model_awd_weekly_forecast)








awdmodel <- auto.arima(awd_xts,ic = "aic",trace = TRUE)
awdmodel2 <- auto.arima(awd_xts2,ic = "aic",trace = TRUE)

awdmodel


```

## Auto-correlation on residuals
```{r}

acf(ts(awdmodel$residuals))
pacf(ts(awdmodel$residuals))
acf(ts(awdmodel2$residuals))
pacf(ts(awdmodel2$residuals))

```

##


```{r}

awd_f =  forecast(awdmodel,level=c(95),h=2*52)
awd_f =  forecast(awdmodel2,level=c(95),h=2*52)
plot(awd_f)

```

```{r time_tk}

awd <- load_awd()

awd_chirps <- bind_rows(awd,chirps_daily_long %>% ungroup()) %>% 
  select(date, parameter, value)



awd_chirps2 <- bind_rows(
  awd2 %>% 
    select(date,parameter, value=pct_change),
  chirps_daily_long %>%
    ungroup() %>% 
    mutate(
      #pct change not so good when 0s
      value= value-lag(value)
    ) %>% 
    filter(!is.na(value))
  ) %>% 
  select(date, parameter, value)
awd_chirps %>% 
  timetk::plot_time_series(.date_var = date,.value = value,.facet_vars = parameter)
awd_chirps2 %>% filter(parameter=="precip")
awd_chirps2 %>% 
  timetk::plot_time_series(.date_var = date,.value = value,.facet_vars = parameter)

awd_chirps %>% count(parameter)
awd_chirps %>%
  group_by(parameter) %>% 
  timetk::plot_acf_diagnostics(.date_var = date,
                       .value = value,
                       .show_white_noise_bars = T)
awd_chirps2 %>%
  group_by(parameter) %>% 
  timetk::plot_acf_diagnostics(.date_var = date,
                       .value = value,
                       .show_white_noise_bars = T)


scale_it <-  function(x){
  (x-mean(x,na.rm=T))/sd(x,na.rm = T)
}

awd_chirps <- awd_chirps %>%
  group_by(parameter) %>% 
  mutate(
    value_scale = scale_it(value)
  ) %>% 
  ungroup()

awd_chirps %>% 
  select(date, parameter, value_scale) %>% 
  pivot_wider(names_from= parameter,values_from = value_scale) %>% 
  timetk::plot_acf_diagnostics(.date_var = date,
                       .value =awd_cases ,
                       .ccf_vars = precip,
                       .show_ccf_vars_only = T)

```



```{r}
awd %>% 
  group_by(yr=lubridate::year(date)) %>% 
  count()
awd <-  awd %>% 
  mutate(uid=row_number())
awd_ts <- ts(awd$value, start=min(awd$uid),end=max(awd$uid),frequency = 52)
class(awd_ts)
library(forecast)
plot(awd_ts)
 plot(awd_xts)


```

https://www.youtube.com/watch?v=dKDgihvtZGk

```{r}




awd_chirps <- bind_rows(chirps_daily_long,awd)
awd_chirps %>% 
  group_by(parameter) %>% 
  summarise(
    min_date= min(date),
    max_date= max(date),
  )

awd_chirps_gte2017 <- awd_chirps %>% 
  ungroup() %>% 
  filter(date>="2017-01-08",date<="2021-09-19")

awd_chirps_gte2017_long <- awd_chirps_gte2017 %>% 
  mutate(
    wk=lubridate::week(date),
    # mo= lubridate::month(date),
    yr = lubridate::year(date)
  ) %>%
  group_by(
    parameter,yr,wk
  ) %>% 
  summarise(
    value= sum(value,na.rm=T), .groups="drop"
  ) 


awd_chirps_gte2017_wide <- awd_chirps_gte2017_long %>% 
  pivot_wider(names_from=parameter, values_from = value) %>% 
  arrange(yr,wk)

rain.ts = ts(awd_chirps_gte2017_wide$precip,
             start = min(awd_chirps_gte2017_wide$wk),
             end = max(awd_chirps_gte2017_wide$wk), frequency=52
             )
awd_chirps_gte2017_wide %>% 
  filter(is.na(awd_cases))
awd_chirps_gte2017_wide <- awd_chirps_gte2017_wide %>% 
  mutate(
    awd_cases= replace_na(awd_cases,0),
    awd_cases_sq = awd_cases^2
  )

awd.ts <- ts(awd_chirps_gte2017_wide$awd_cases,
             start = min(awd_chirps_gte2017_wide$wk),
             end = max(awd_chirps_gte2017_wide$wk),
             frequency = 52)

# error at ccf point in there.
# debugonce(astsa::lag2.plot)

windows();astsa::lag2.plot(rain.ts,awd.ts,52,max.lag = 20)

awd_chirps_gte2017_wide2 <- awd_chirps_gte2017_wide %>% 
  mutate(
    awd_diff = awd_cases-lag(awd_cases)
    # awd_diff = (awd_cases-lag(awd_cases))/lag(awd_cases)
    
    ) %>% 
  filter(!is.na(awd_diff))
# awd_chirps_gte2017_wide2 <- awd_chirps_gte2017_wide %>% 
#   filter(yr!=2017 & wk!=2)
awd_chirps_gte2017_wide2 %>% 
  ggplot(aes(x=awd_cases, y= precip))+
  geom_point()+
  scale_x_log10()
awd_chirps_gte2017_wide2 %>% 
  ggplot(aes(x=precip))+
  geom_histogram()+
  scale_x_log10()


awd.ts2 <- ts(awd_chirps_gte2017_wide2$awd_diff,
             start = min(awd_chirps_gte2017_wide2$wk),
             end = max(awd_chirps_gte2017_wide2$wk),
             frequency = 52)

rain.ts2 = ts(awd_chirps_gte2017_wide2$precip,
             start = min(awd_chirps_gte2017_wide2$wk),
             end = max(awd_chirps_gte2017_wide2$wk), frequency=52
             )

windows();astsa::lag2.plot(rain.ts2,
                           awd.ts2,
                           52,
                           max.lag = 20)



ck <- awd_chirps_gte2017 %>% 
  group_by(parameter,yr= lubridate::year(date),mo=lubridate::month(date)) %>% 
  summarise(value=sum(value),.groups="drop") %>% 
  mutate(date= lubridate::ymd(glue("{yr}-{mo}-01"))) %>% 
  select(date,parameter,value) %>% 
  pivot_wider(names_from = parameter, values_from = value) 
ck %>% 
  ggplot(aes(x=date,y=awd_cases))+
  geom_line()+
  geom_line(data=ck, aes(x=date,y= precip*10))

ck %>% 
  ggplot(aes(x=precip, y= awd_cases))+
  geom_point()+
  scale_y_log10()
```